---
title: "Measles, Bangladesh 2026: descriptive epidemiology of public data"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: false
    fig-width: 9
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---

Specification: `docs/describe-spec.md`.
Public data only.

::: {.callout-note collapse="true"}
## How this report is written

- Titles state what, where and when. Source and units are in each caption.
- Collapsed "Scaffold" boxes hold the numbers for each display's descriptive text, still to be written.
- Deaths are shown under two readings until DGHS confirms how they count (`who-questions.md` Q6). A: suspected deaths include confirmed deaths. B: suspected and confirmed deaths are separate, so total = suspected + confirmed.
- Place is the place of the reporting facility, not residence.
:::

# Setup and inputs

```{r}
#| label: setup

library(here)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(plotly)
library(sf)
library(slider)
library(knitr)

# visualise-data palette
col_primary <- "#1a1a1a"
col_highlight <- "#c0392b"
col_reference <- "#7f7f7f"
theme_set(
  theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank(), legend.position = "bottom")
)
sf_use_s2(FALSE)

source_dghs <- "Source: DGHS measles monitoring platform (measles.dghs.gov.bd), report date."
```

## Case series

The dashboard series by report date.
2 to 5 April are empty, and 7 April holds everything from 15 March to 7 April (`flag`), so daily series start on 8 April.
Cumulative counts include the 7 April lump.

```{r}
#| label: inputs-cases

cases <- read.csv(here("data", "dghs-cases.csv"), na.strings = c("", "NA")) |>
  mutate(date = ymd(date))
cutoff <- max(cases$date)
first_daily <- ymd("2026-04-08")

# DGHS district spellings that differ from the OCHA boundaries
district_recode <- c(
  Jhalakathi = "Jhalokati",
  Netrokona = "Netrakona",
  Chapainawabganj = "Chapainababganj"
)
cases <- cases |>
  mutate(district = recode(district, !!!district_recode))
```

Data cutoff: `r format(cutoff, "%d %B %Y")`.

## Geography and population

Divisions are ordered west to east by the longitude of their centroid, and districts the same way within each division.
Population is the 2022 census projection (`cod-ps-bgd`), not adjusted for growth.

```{r}
#| label: inputs-geography

adm1 <- st_read(here("data", "covariates", "bgd_adm1.geojson"), quiet = TRUE)
adm2 <- st_read(here("data", "covariates", "bgd_adm2.geojson"), quiet = TRUE)
lookup <- read.csv(here("data", "covariates", "admin_lookup.csv"))

centroid_lon <- function(g) {
  suppressWarnings(st_coordinates(st_centroid(g))[, 1])
}
division_order <- adm1 |>
  mutate(lon = centroid_lon(geometry)) |>
  arrange(lon) |>
  pull(adm1_name)
district_order <- adm2 |>
  mutate(
    lon = centroid_lon(geometry),
    division = factor(adm1_name, levels = division_order)
  ) |>
  arrange(division, lon) |>
  pull(adm2_name)

pop1 <- read.csv(here("data", "covariates", "population_adm1.csv")) |>
  select(adm1_pcode, pop = t_tl, pop_u5 = t_00_04) |>
  left_join(distinct(lookup, adm1_pcode, division = adm1_name),
            by = "adm1_pcode")
pop2 <- read.csv(here("data", "covariates", "population_adm2.csv")) |>
  select(adm2_pcode, pop = t_tl, pop_u5 = t_00_04) |>
  left_join(select(lookup, adm2_pcode, district = adm2_name,
                   division = adm1_name),
            by = "adm2_pcode")

stopifnot(
  setequal(pop1$division, unique(na.omit(cases$division))),
  setequal(pop2$district, unique(na.omit(cases$district)))
)

division_order
```

## Derived series

```{r}
#| label: inputs-derived

# Wide daily series: one row per date and place, one column per measure
widen <- function(d) {
  d |>
    select(-fetched_at, -flag) |>
    pivot_wider(names_from = measure, values_from = value)
}
div_daily <- cases |>
  filter(level == "division") |>
  widen() |>
  mutate(division = factor(division, levels = division_order))
nat_daily <- cases |>
  filter(level == "national") |>
  widen()
dist_daily <- cases |>
  filter(level == "district") |>
  widen() |>
  mutate(district = factor(district, levels = district_order),
         division = factor(division, levels = division_order))

# Weeks are 7-day blocks ending on the cutoff, so the last week is complete;
# the first, partial block is dropped.
week_end_of <- function(date) {
  cutoff - 7 * floor(as.numeric(cutoff - date) / 7)
}
complete_weeks <- function(d) {
  d |>
    mutate(week_end = week_end_of(date)) |>
    filter(week_end - 6 >= first_daily)
}

# Trailing 7-day sum, plotted at the window's last day
trail7 <- function(x) slide_dbl(x, sum, .before = 6, .complete = TRUE)
```

## Events

```{r}
#| label: inputs-events

events <- read.csv(here("assets", "events.csv")) |>
  mutate(date = ymd(date))
events_marked <- events |>
  filter(!grepl("backfill", event))
kable(events, caption = "Events marked on time series (assets/events.csv)")
```

Not yet marked: campaign rounds by division (`who-questions.md` Q17) and the September mop-up dates.

# What

## Measures

```{r}
#| label: tbl-measures
#| tbl-cap: "DGHS dashboard measures, 24-hour counts by report date, Bangladesh, 2026"

tibble::tribble(
  ~measure, ~dashboard_label, ~notes,
  "suspected", "suspected24h", "Hospital admissions with suspected measles (WHO call); inclusion of confirmed unknown (Q6)",
  "confirmed", "confirmed24h", "Laboratory method not stated; about 1,000 PCR-confirmed in total (WHO call)",
  "suspected_deaths", "suspectedDeath24h", "Inclusion of confirmed deaths unknown (Q6); place of death unknown (Q7)",
  "confirmed_deaths", "confirmedDeath24h", "As above",
  "admitted", "admitted24h", "Relation to suspected not published",
  "discharged", "discharged24h", "\"Recovered\" in the dashboard's Excel export",
  "serum_sent", "serumSent24h", "Serum samples sent to the laboratory; not in the Excel export"
) |>
  kable()
```

```{r}
#| label: tbl-who-definitions
#| tbl-cap: "WHO measles surveillance case definitions, used in WHO monthly data (verbatim)"

read.csv(here("assets", "case-definitions.csv")) |>
  select(classification, definition, page) |>
  kable()
```

No official Bangladesh 2026 case definition was found in public documents.

## Laboratory testing in the DGHS series

Share sampled is serum samples sent over suspected cases.
Confirmation yield is confirmed cases over serum samples sent, in the same week (lag 0) and the week before (lag 1 week).
Yield is not test positivity: confirmations lag samples, and not every confirmation need come from a serum sample.

```{r}
#| label: fig-testing
#| fig-cap: "Serum samples sent per suspected case, and confirmed cases per serum sample sent, by division and week, Bangladesh, May 2026 to cutoff. Source: DGHS measles monitoring platform, report date."
#| fig-alt: "Line charts, one panel per division, of weekly testing ratios over time."
#| fig-height: 7

# Serum samples are recorded from May (R/data/01-fetch-dashboard.R), so the
# ratios start with the first complete week in May.
serum_start <- ymd("2026-05-01")
testing <- div_daily |>
  complete_weeks() |>
  filter(week_end - 6 >= serum_start) |>
  group_by(division, week_end) |>
  summarise(across(c(suspected, confirmed, serum_sent), sum),
            .groups = "drop") |>
  arrange(division, week_end) |>
  group_by(division) |>
  mutate(
    share_sampled = serum_sent / suspected,
    yield_lag0 = if_else(serum_sent > 0, confirmed / serum_sent, NA),
    yield_lag1 = if_else(lag(serum_sent) > 0,
                         confirmed / lag(serum_sent), NA)
  ) |>
  ungroup()

p <- testing |>
  pivot_longer(c(share_sampled, yield_lag0, yield_lag1),
               names_to = "ratio") |>
  mutate(ratio = recode(ratio,
                        share_sampled = "Serum sent / suspected",
                        yield_lag0 = "Confirmed / serum sent, same week",
                        yield_lag1 = "Confirmed / serum sent, week before")) |>
  ggplot(aes(week_end, value, colour = ratio, linetype = ratio)) +
  geom_line() +
  facet_wrap(~division, ncol = 4, scales = "free_y") +
  scale_colour_manual(values = c(
    "Serum sent / suspected" = col_primary,
    "Confirmed / serum sent, same week" = col_highlight,
    "Confirmed / serum sent, week before" = col_highlight)) +
  scale_linetype_manual(values = c(
    "Serum sent / suspected" = "solid",
    "Confirmed / serum sent, same week" = "solid",
    "Confirmed / serum sent, week before" = "dashed")) +
  scale_x_date(date_labels = "%b") +
  labs(x = "Week ending", y = "Ratio", colour = NULL, linetype = NULL)
ggplotly(p) |> layout(legend = list(orientation = "h", y = -0.15))
```

::: {.callout-tip collapse="true"}
## Scaffold: testing

- Share sampled, national, from May: `r with(nat_daily |> filter(date >= serum_start), round(sum(serum_sent) / sum(suspected), 2))`.
- Confirmed / serum sent, national, from May: `r with(nat_daily |> filter(date >= serum_start), round(sum(confirmed) / sum(serum_sent), 2))`.
- Range of weekly share sampled across divisions: `r paste(round(range(testing$share_sampled, na.rm = TRUE), 2), collapse = " to ")`.
:::

```{r}
#| label: checks-testing

# Reported, not enforced: confirmations need not come from serum samples
# recorded on the platform.
cum_div <- div_daily |>
  filter(date >= serum_start) |>
  group_by(division) |>
  summarise(across(c(suspected, confirmed, serum_sent), sum)) |>
  mutate(confirmed_le_suspected = confirmed <= suspected,
         serum_ge_confirmed = serum_sent >= confirmed)
stopifnot(all(cum_div$confirmed_le_suspected))
kable(cum_div, caption = "Cumulative counts from 1 May, by division")
```

## Case classification in WHO monthly surveillance

WHO monthly data come from EPI case-based surveillance, a different stream from the DGHS hospital series.
Share tested = (laboratory-confirmed + discarded) / suspected.
Laboratory positivity = laboratory-confirmed / (laboratory-confirmed + discarded).

```{r}
#| label: fig-who-classification
#| fig-cap: "Measles cases by final classification, and share of suspected cases tested and laboratory positivity, by month, Bangladesh, January 2023 to latest. Source: WHO provisional monthly measles and rubella data (table 404)."
#| fig-alt: "Stacked bar chart of monthly cases by classification above a line chart of share tested and positivity."
#| fig-height: 8

who <- read.csv(here("data", "who-monthly.csv")) |>
  mutate(month = make_date(year, month, 1)) |>
  filter(month >= ymd("2023-01-01"))
stopifnot(all(who$measles_suspect == who$measles_total + who$discarded))

who_long <- who |>
  select(month, `Laboratory-confirmed` = measles_lab_confirmed,
         `Epidemiologically linked` = measles_epi_linked,
         `Clinically compatible` = measles_clinical,
         Discarded = discarded) |>
  pivot_longer(-month, names_to = "classification", values_to = "cases") |>
  mutate(classification = factor(classification, levels = c(
    "Discarded", "Clinically compatible", "Epidemiologically linked",
    "Laboratory-confirmed")))

p_counts <- ggplot(who_long, aes(month, cases, fill = classification)) +
  geom_col(width = 25) +
  scale_fill_manual(values = c("#d4d4d4", col_reference, col_highlight,
                               col_primary)) +
  scale_x_date(date_labels = "%b %Y") +
  labs(x = NULL, y = "Suspected cases", fill = NULL)

who_ratios <- who |>
  mutate(
    tested = measles_lab_confirmed + discarded,
    `Share tested` = tested / measles_suspect,
    `Laboratory positivity` = if_else(tested > 0,
                                      measles_lab_confirmed / tested, NA)
  ) |>
  select(month, `Share tested`, `Laboratory positivity`) |>
  pivot_longer(-month, names_to = "ratio")
p_ratios <- ggplot(who_ratios, aes(month, value, colour = ratio)) +
  geom_line() +
  geom_point(size = 1) +
  scale_colour_manual(values = c(col_highlight, col_primary)) +
  scale_x_date(date_labels = "%b %Y") +
  labs(x = NULL, y = "Proportion", colour = NULL)

subplot(ggplotly(p_counts), ggplotly(p_ratios), nrows = 2, shareX = TRUE,
        titleY = TRUE, heights = c(0.6, 0.4))
```

In every month, suspected = total + discarded.
Discarded cases include those discarded without a laboratory test (WHO definition, `tbl-who-definitions`), so share tested is an upper bound.

## Data quality

```{r}
#| label: tbl-data-quality
#| tbl-cap: "Failed checks and flagged dates, DGHS dashboard series, Bangladesh, 2026"

checks <- read.csv(here("data", "dghs-cases-checks.csv"))
failed <- checks |>
  filter(!pass) |>
  transmute(item = check, date, division, measure,
            detail = paste("expected", expected, "observed", observed))
flagged <- cases |>
  filter(!is.na(flag), level == "national", measure == "suspected") |>
  transmute(item = flag, date = as.character(date), division = NA,
            measure = NA, detail = paste("national suspected", value))
bind_rows(failed, flagged) |> kable()
```

```{r}
#| label: tbl-source-agreement
#| tbl-cap: "Agreement of dashboard and press release counts by month, Bangladesh, 2026"

read.csv(here("sitrep", "data", "dghs-source-compare.csv")) |>
  mutate(month = format(ymd(date), "%Y-%m")) |>
  filter(!is.na(diff)) |>
  group_by(month) |>
  summarise(comparisons = n(), differ = sum(diff != 0),
            net_difference = sum(diff)) |>
  kable()
```

## Reporting units

Reporting units that submitted each day, by division.
Outside Dhaka division, units are district civil surgeon offices plus, in some divisions, one other unit.
The number of registered units is not published (`who-questions.md` Q21), so the expected number is taken as the most frequent daily count in each division, and completeness is reports over that number.
Dhaka division varies too widely for this and is shown as a count.

```{r}
#| label: fig-reporting-units
#| fig-cap: "Reporting units submitting per day, by division, Bangladesh, 8 April 2026 to cutoff. Divisions other than Dhaka: reports as a proportion of the most frequent daily count. Dhaka: count of reports. Source: DGHS measles monitoring platform."
#| fig-alt: "Line charts of daily reporting completeness by division and the daily count of reports in Dhaka division."
#| fig-height: 7

n_districts <- lookup |> count(division = adm1_name, name = "districts")
modal <- function(x) as.integer(names(which.max(table(x))))
reports <- div_daily |>
  filter(date >= first_daily) |>
  select(date, division, report_count) |>
  group_by(division) |>
  mutate(expected = modal(report_count)) |>
  ungroup() |>
  left_join(n_districts, by = "division") |>
  mutate(value = if_else(division == "Dhaka", report_count,
                         report_count / expected))
p <- ggplot(reports, aes(date, value)) +
  geom_line(colour = col_primary) +
  facet_wrap(~division, ncol = 4, scales = "free_y") +
  scale_x_date(date_labels = "%b") +
  labs(x = "Report date", y = "Reports / expected (Dhaka: count)")
ggplotly(p)

reports |>
  distinct(division, districts, expected) |>
  kable(caption = "Districts and most frequent daily count of reports, by division")
```

# How much

```{r}
#| label: tbl-cumulative
#| tbl-cap: "Cumulative counts and suspected cases per 100,000 population, by division, Bangladesh, 15 March 2026 to cutoff. Source: DGHS measles monitoring platform; population: 2022 census projection."

u5_national <- sum(pop1$pop_u5) / sum(pop1$pop)

cumulate <- function(d) {
  d |>
    summarise(across(c(suspected, confirmed, suspected_deaths,
                       confirmed_deaths, admitted, discharged), sum))
}
coxs_pcode <- lookup$adm2_pcode[lookup$adm2_name == "Cox's Bazar"]
coxs_pop <- pop2$pop[pop2$adm2_pcode == coxs_pcode]
chattogram_pop <- pop1$pop[pop1$division == "Chattogram"]

cum_table <- bind_rows(
  div_daily |>
    group_by(place = as.character(division)) |>
    cumulate() |>
    left_join(pop1, by = c("place" = "division")),
  nat_daily |>
    cumulate() |>
    mutate(place = "National", pop = sum(pop1$pop),
           pop_u5 = sum(pop1$pop_u5)),
  dist_daily |>
    filter(district == "Cox's Bazar") |>
    cumulate() |>
    mutate(place = "Cox's Bazar district", pop = coxs_pop,
           pop_u5 = pop2$pop_u5[pop2$adm2_pcode == coxs_pcode]),
  dist_daily |>
    filter(division == "Chattogram", district != "Cox's Bazar") |>
    cumulate() |>
    mutate(place = "Chattogram excluding Cox's Bazar",
           pop = chattogram_pop - coxs_pop,
           pop_u5 = pop1$pop_u5[pop1$division == "Chattogram"] -
             pop2$pop_u5[pop2$adm2_pcode == coxs_pcode])
) |>
  mutate(
    place = factor(place, levels = c(division_order, "National",
                                     "Cox's Bazar district",
                                     "Chattogram excluding Cox's Bazar")),
    suspected_per_100k = 1e5 * suspected / pop,
    u5_share_pct = 100 * pop_u5 / pop,
    u5_share_band = case_when(
      is.na(u5_share_pct) | place == "National" ~ NA,
      u5_share_pct > 100 * u5_national ~ "above national",
      TRUE ~ "below national"
    )
  ) |>
  arrange(place)

cum_table |>
  select(place, suspected, confirmed, suspected_deaths, confirmed_deaths,
         suspected_per_100k, u5_share_pct, u5_share_band) |>
  kable(digits = 1, format.args = list(big.mark = ","))
```

Under-five share of population: national `r round(100 * u5_national, 1)`%.
No under-five rates: public data have no cases by age and division.
Whether the census counts camp residents is unknown (Q12).

```{r}
#| label: fig-incidence
#| fig-cap: "Suspected cases per 100,000 population in the previous 7 days, by division (black) with national (grey), Bangladesh, 14 April 2026 to cutoff. Log scale. Source: DGHS measles monitoring platform; population: 2022 census projection."
#| fig-alt: "Line charts, one panel per division, of 7-day incidence per 100,000 on a log scale, each with the national series in grey."
#| fig-height: 6

incidence <- div_daily |>
  filter(date >= first_daily) |>
  arrange(division, date) |>
  group_by(division) |>
  mutate(cases_7d = trail7(suspected)) |>
  ungroup() |>
  left_join(pop1, by = "division") |>
  mutate(incidence_7d = 1e5 * cases_7d / pop) |>
  filter(!is.na(incidence_7d))
incidence_nat <- nat_daily |>
  filter(date >= first_daily) |>
  arrange(date) |>
  mutate(incidence_7d = 1e5 * trail7(suspected) / sum(pop1$pop)) |>
  filter(!is.na(incidence_7d))

p <- ggplot(incidence, aes(date, incidence_7d)) +
  geom_line(data = select(incidence_nat, date, incidence_7d),
            colour = col_reference) +
  geom_line(colour = col_primary) +
  facet_wrap(~division, ncol = 4) +
  scale_y_log10() +
  scale_x_date(date_labels = "%b") +
  labs(x = "Report date", y = "Suspected cases per 100,000, previous 7 days")
ggplotly(p)
```

```{r}
#| label: checks-how-much

stopifnot(
  nrow(pop1) == 8, nrow(pop2) == 64,
  isTRUE(all.equal(sum(cum_table$suspected[cum_table$place %in% division_order]),
                   cum_table$suspected[cum_table$place == "National"]))
)
```

# When

## National timeline

```{r}
#| label: fig-timeline
#| fig-cap: "Measles cases per month: WHO surveillance suspected and total cases (bars), and DGHS hospital suspected cases (points), Bangladesh, January 2026 to cutoff. DGHS April is 8 to 30 April; 15 March to 7 April is reported as one total; the current month is shown when complete. Sources: WHO provisional monthly measles and rubella data; DGHS measles monitoring platform."
#| fig-alt: "Monthly bar chart of WHO suspected and total measles cases with DGHS monthly suspected hospital cases as points."

who_2026 <- who |>
  filter(year(month) == 2026) |>
  select(month, `WHO suspected` = measles_suspect,
         `WHO total (lab + epi-linked + clinical)` = measles_total) |>
  pivot_longer(-month, names_to = "series", values_to = "cases")
dghs_monthly <- nat_daily |>
  filter(date >= first_daily) |>
  group_by(month = floor_date(date, "month")) |>
  summarise(cases = sum(suspected), days = n(), .groups = "drop") |>
  # Current month dropped while incomplete; April starts on 8 April
  filter(month < floor_date(cutoff + 1, "month")) |>
  mutate(series = "DGHS suspected (hospital)")
lump <- nat_daily |> filter(date == ymd("2026-04-07")) |> pull(suspected)

p <- ggplot() +
  geom_col(data = who_2026,
           aes(month, cases, fill = series),
           position = position_dodge(width = 25), width = 24) +
  geom_point(data = dghs_monthly, aes(month, cases, text = paste(days, "days")),
             colour = col_highlight, size = 2.5) +
  geom_vline(xintercept = ymd("2026-06-01"), linetype = "dotted",
             colour = col_reference) +
  annotate("text", x = ymd("2026-06-01"), y = max(dghs_monthly$cases),
           label = "WHO: classification change (June)", hjust = -0.05,
           size = 3, colour = col_reference) +
  scale_fill_manual(values = c(col_reference, "#d4d4d4")) +
  scale_x_date(date_labels = "%b") +
  labs(x = NULL, y = "Cases per month", fill = NULL)
ggplotly(p, tooltip = c("x", "y", "text"))
```

DGHS cases from 15 March to 7 April, reported as one total on 7 April: `r format(lump, big.mark = ",")`.

## Daily cases by division

```{r}
#| label: fig-epicurve-division
#| fig-cap: "Daily suspected cases (points; Saturdays red) and 7-day trailing mean (line), by division, Bangladesh, 8 April 2026 to cutoff. Log scale; days with zero cases not shown. Vertical lines: events in assets/events.csv. Crosses at the foot of a panel: days with fewer reports than the division's most frequent daily count (not shown for Dhaka). Source: DGHS measles monitoring platform, report date."
#| fig-alt: "Daily epidemic curves on a log scale, one panel per division, with 7-day means, campaign start lines and markers for days with missing district reports."
#| fig-height: 7

epicurve <- div_daily |>
  filter(date >= first_daily) |>
  arrange(division, date) |>
  group_by(division) |>
  mutate(mean_7d = trail7(suspected) / 7) |>
  ungroup() |>
  left_join(distinct(reports, division, expected), by = "division") |>
  mutate(saturday = wday(date, week_start = 1) == 6,
         missing_report = division != "Dhaka" & report_count < expected)
n_zero <- sum(epicurve$suspected == 0)

p <- ggplot(epicurve, aes(date)) +
  geom_vline(data = events_marked, aes(xintercept = date),
             colour = col_reference, linetype = "dotted") +
  geom_point(data = filter(epicurve, suspected > 0),
             aes(y = suspected, colour = saturday), size = 0.7) +
  geom_line(aes(y = mean_7d), colour = col_primary) +
  geom_point(data = filter(epicurve, missing_report), aes(y = 1),
             shape = 4, colour = col_highlight) +
  facet_wrap(~division, ncol = 4) +
  scale_colour_manual(values = c(`FALSE` = col_reference,
                                 `TRUE` = col_highlight), guide = "none") +
  scale_y_log10() +
  scale_x_date(date_labels = "%b") +
  labs(x = "Report date", y = "Suspected cases per day")
ggplotly(p)
```

Division-days with zero suspected cases (not shown): `r n_zero`.
Division-days with fewer reports than expected: `r sum(epicurve$missing_report)`.

## Timing by division

```{r}
#| label: tbl-timing
#| tbl-cap: "Peak and recent change in daily suspected cases (7-day trailing mean), by division, Bangladesh, 8 April 2026 to cutoff. Source: DGHS measles monitoring platform, report date."

weekly_div <- div_daily |>
  complete_weeks() |>
  group_by(division, week_end) |>
  summarise(suspected = sum(suspected), .groups = "drop") |>
  arrange(division, week_end) |>
  group_by(division) |>
  mutate(ratio = suspected / lag(suspected)) |>
  slice_tail(n = 4) |>
  mutate(k = paste0("ratio_wk", row_number())) |>
  ungroup()

timing <- epicurve |>
  filter(!is.na(mean_7d)) |>
  group_by(division) |>
  summarise(
    peak_date = date[which.max(mean_7d)],
    peak_mean = max(mean_7d),
    current_mean = last(mean_7d),
    current_pct_of_peak = 100 * current_mean / peak_mean,
    peak_censored = peak_date %in% range(date)
  ) |>
  left_join(
    weekly_div |>
      select(division, k, ratio) |>
      pivot_wider(names_from = k, values_from = ratio),
    by = "division"
  ) |>
  mutate(
    # Doubling (positive) or halving (negative) time in days, last week
    doubling_days = 7 * log(2) / log(ratio_wk4)
  )
kable(timing, digits = 2)
```

Week-on-week ratios are for the four 7-day blocks ending `r format(cutoff, "%d %B")` (`ratio_wk4` is the latest).
Negative doubling times are halving times.

## Day of week

```{r}
#| label: tbl-weekday
#| tbl-cap: "Daily suspected cases relative to the centred 7-day mean, by weekday and division: median (interquartile range), Bangladesh, 8 April 2026 to cutoff. Source: DGHS measles monitoring platform, report date."

weekday <- div_daily |>
  filter(date >= first_daily) |>
  arrange(division, date) |>
  group_by(division) |>
  mutate(centred = slide_dbl(suspected, mean, .before = 3, .after = 3,
                             .complete = TRUE),
         rel = suspected / centred) |>
  ungroup() |>
  filter(!is.na(rel)) |>
  mutate(weekday = wday(date, label = TRUE, week_start = 6)) |>
  group_by(division, weekday) |>
  summarise(cell = sprintf("%.2f (%.2f to %.2f)", median(rel),
                           quantile(rel, 0.25), quantile(rel, 0.75)),
            .groups = "drop") |>
  pivot_wider(names_from = weekday, values_from = cell)
kable(weekday)
```

Weeks start on Saturday, the first day of the Bangladesh working week.

## Intervals that can be estimated

```{r}
#| label: tbl-timespans
#| tbl-cap: "Intervals in the 2026 Bangladesh measles outbreak and the data that can estimate each"

tibble::tribble(
  ~interval, ~public_data, ~who_line_list, ~other_source,
  "Exposure to onset (incubation)", "No", "No", "Literature (parameter register)",
  "Generation or serial interval", "No", "No (no transmission pairs)", "Literature",
  "Fever to rash", "No", "Yes (DOnsetF, DOnsetR)", "",
  "Rash to notification", "No", "Yes (DOnsetR, DNOT)", "",
  "Notification to investigation", "No", "Yes (DNOT, DOI)", "",
  "Rash to specimen; specimen to result", "No", "Yes (lab dates, about half of records)", "",
  "Onset to admission", "No", "Only if admission date is digitised (Q2)", "",
  "Admission to report date", "No; indirect signs only (days with missing reports)", "No", "WHO or DGHS (Q22)",
  "Suspected to confirmed", "No (share tested varies)", "Yes", "",
  "Admission to exit (discharge or death)", "Partly: deconvolution of discharges and deaths against admissions, if deaths are hospital deaths (Q7)", "Only if outcome fields are digitised (Q2)", "",
  "Onset to death", "No", "Free-text Comment only", "Death line list (Q2)",
  "Campaign dose to rash", "No", "DateLastMCV, 0.5% complete", "",
  "Undetected period (first infection to detection)", "No", "Onset dates before April, if shared", "Molecular clock on sequences (IEDCR)"
) |>
  kable(col.names = c("Interval", "Public data", "WHO line list",
                      "Other source"))
```

# Where

```{r}
#| label: inputs-where

district_cum <- dist_daily |>
  group_by(district) |>
  summarise(suspected = sum(suspected), .groups = "drop") |>
  left_join(pop2, by = "district") |>
  mutate(per_100k = 1e5 * suspected / pop)
stopifnot(nrow(district_cum) == 64, !anyNA(district_cum$pop))

map_districts <- adm2 |>
  left_join(district_cum, by = c("adm2_name" = "district"))
stopifnot(!anyNA(map_districts$per_100k))

medical_colleges <- read.csv(here("data", "covariates", "hospitals.csv")) |>
  filter(medical_college) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)
breaks_100k <- c(0, 25, 50, 100, 200, 400)
stopifnot(max(map_districts$per_100k) <= max(breaks_100k))
```

```{r}
#| label: fig-map-cumulative
#| fig-cap: "Cumulative suspected cases per 100,000 population, by district of reporting facility, Bangladesh, 15 March 2026 to cutoff. Points: hospitals named \"medical college\" in OpenStreetMap (unchecked list; referral hospitals). Sources: DGHS measles monitoring platform; 2022 census projection; Healthsites.io, © OpenStreetMap contributors."
#| fig-alt: "Choropleth map of Bangladesh districts shaded by cumulative suspected measles cases per 100,000, with division outlines and points for medical college hospitals."
#| fig-height: 8

ggplot() +
  geom_sf(data = map_districts, aes(fill = per_100k), colour = "white",
          linewidth = 0.1) +
  geom_sf(data = adm1, fill = NA, colour = col_primary, linewidth = 0.4) +
  geom_sf(data = medical_colleges, shape = 21, size = 1, fill = "white",
          colour = col_primary) +
  scale_fill_viridis_b(breaks = breaks_100k, limits = range(breaks_100k),
                       option = "magma", direction = -1,
                       show.limits = TRUE) +
  labs(fill = "Suspected cases\nper 100,000") +
  theme_void() +
  theme(legend.position = "right")
```

```{r}
#| label: fig-district-heatmap
#| fig-cap: "Suspected cases per 100,000 population per week, by district of reporting facility, Bangladesh, weeks ending 14 April 2026 to cutoff. Districts grouped by division, west to east. Source: DGHS measles monitoring platform; 2022 census projection."
#| fig-alt: "Heat map with one row per district and one column per week, shaded by weekly suspected cases per 100,000."
#| fig-height: 13

heat <- dist_daily |>
  complete_weeks() |>
  group_by(division, district, week_end) |>
  summarise(suspected = sum(suspected), .groups = "drop") |>
  left_join(select(pop2, district, pop), by = "district") |>
  mutate(per_100k = 1e5 * suspected / pop,
         row = factor(paste0(district, " (", division, ")"),
                      levels = rev(paste0(district_order, " (",
                                          pop2$division[match(district_order,
                                                              pop2$district)],
                                          ")"))))
p <- ggplot(heat, aes(week_end, row, fill = per_100k,
                      text = paste0(district, ", week ending ", week_end,
                                    ": ", round(per_100k, 1), " per 100,000 (",
                                    suspected, " cases)"))) +
  geom_tile() +
  scale_fill_viridis_b(breaks = c(0, 2, 5, 10, 20, 40), option = "magma",
                       direction = -1) +
  scale_x_date(date_labels = "%b", expand = c(0, 0)) +
  labs(x = "Week ending", y = NULL, fill = "Per 100,000\nper week") +
  theme(panel.grid = element_blank())
ggplotly(p, tooltip = "text", height = 1100)
```

```{r}
#| label: fig-map-peak
#| fig-cap: "Month of the highest 7-day mean of suspected cases, by district of reporting facility, Bangladesh, 8 April 2026 to cutoff. Red outline: peak in the first or last week of the series. Source: DGHS measles monitoring platform."
#| fig-alt: "Choropleth map of districts coloured by the month in which suspected cases peaked."
#| fig-height: 8

peaks <- dist_daily |>
  filter(date >= first_daily) |>
  arrange(district, date) |>
  group_by(district) |>
  mutate(mean_7d = trail7(suspected) / 7) |>
  filter(!is.na(mean_7d)) |>
  summarise(peak_date = date[which.max(mean_7d)],
            first = min(date), last = max(date), .groups = "drop") |>
  mutate(peak_month = factor(format(peak_date, "%B"),
                             levels = month.name[4:10]),
         censored = peak_date - first < 7 | last - peak_date < 7)

map_peaks <- adm2 |> left_join(peaks, by = c("adm2_name" = "district"))
ggplot() +
  geom_sf(data = map_peaks, aes(fill = peak_month), colour = "white",
          linewidth = 0.1) +
  geom_sf(data = filter(map_peaks, censored), fill = NA,
          colour = col_highlight, linewidth = 0.5) +
  geom_sf(data = adm1, fill = NA, colour = col_primary, linewidth = 0.4) +
  scale_fill_viridis_d(option = "viridis", drop = TRUE) +
  labs(fill = "Peak month") +
  theme_void() +
  theme(legend.position = "right")
```

Districts whose peak falls in the first or last week (red outline): `r paste(sort(peaks$district[peaks$censored]), collapse = ", ")`.

```{r}
#| label: fig-map-hospitals
#| fig-cap: "Hospitals per 100,000 population in OpenStreetMap, by district, Bangladesh, 2026. Context only: mapping completeness varies by district. Sources: Healthsites.io, © OpenStreetMap contributors; 2022 census projection."
#| fig-alt: "Choropleth map of OpenStreetMap hospitals per 100,000 population by district."
#| fig-height: 8

hospitals_district <- read.csv(here("data", "covariates", "hospitals.csv")) |>
  count(adm2_pcode, name = "hospitals")
map_hospitals <- adm2 |>
  left_join(hospitals_district, by = "adm2_pcode") |>
  left_join(select(pop2, adm2_pcode, pop), by = "adm2_pcode") |>
  mutate(hospitals = coalesce(hospitals, 0L),
         per_100k = 1e5 * hospitals / pop)
ggplot() +
  geom_sf(data = map_hospitals, aes(fill = per_100k), colour = "white",
          linewidth = 0.1) +
  geom_sf(data = adm1, fill = NA, colour = col_primary, linewidth = 0.4) +
  scale_fill_viridis_b(option = "mako", direction = -1, n.breaks = 6) +
  labs(fill = "Hospitals\nper 100,000") +
  theme_void() +
  theme(legend.position = "right")
```

Campaign coverage by division and city corporation (`fig-campaign-coverage`) waits on `sitrep/data/dghs-campaign.csv`.

# Among whom, including camps

```{r}
#| label: inputs-among-whom

read_confirmed <- function(path) {
  if (!file.exists(path)) return(NULL)
  d <- read.csv(path, colClasses = c(confirmed_by = "character"))
  filter(d, !is.na(confirmed_by), confirmed_by != "")
}
context_figures <- read_confirmed(here("data", "context-figures.csv"))
camps_measles <- read_confirmed(here("data", "camps-measles.csv"))
n_confirmed <- c(context = NROW(context_figures), camps = NROW(camps_measles))
n_confirmed
```

Age, sex and vaccination figures (`tbl-reported-figures`, `fig-age-shares`) and camp figures (`tbl-camps`, `fig-camps`) show only rows a person has confirmed in `data/context-figures.csv` and `data/camps-measles.csv`.

```{r}
#| label: tbl-reported-figures
#| tbl-cap: "Reported measles case figures by age, sex and vaccination status, from WHO, UN and UNICEF documents, Bangladesh, 2026"

if (n_confirmed[["context"]] > 0) {
  context_figures |>
    select(source, doc_id, published_date, page, indicator, value, unit,
           denominator, geography, quote) |>
    kable()
} else {
  cat("No confirmed rows yet.")
}
```

```{r}
#| label: tbl-camp-population
#| tbl-cap: "Population of Rohingya camps and Bhasan Char (UNHCR, end 2025) and Cox's Bazar district (2022 census projection)"

camp_pop <- read.csv(here("data", "covariates", "camp_population.csv")) |>
  filter(year == max(year), population_type == "REF",
         grepl("Camp|RC|Bhasan Char", location))
tibble::tibble(
  population = c("Rohingya camps, Cox's Bazar (UNHCR)",
                 "Bhasan Char (UNHCR)",
                 "Cox's Bazar district (census projection)"),
  total = c(sum(camp_pop$t_tl[!grepl("Bhasan", camp_pop$location)]),
            sum(camp_pop$t_tl[grepl("Bhasan", camp_pop$location)]),
            coxs_pop),
  under_five = c(sum(camp_pop$t_00_04[!grepl("Bhasan", camp_pop$location)]),
                 sum(camp_pop$t_00_04[grepl("Bhasan", camp_pop$location)]),
                 pop2$pop_u5[pop2$adm2_pcode == coxs_pcode])
) |>
  mutate(under_five_pct = 100 * under_five / total) |>
  kable(digits = 1, format.args = list(big.mark = ","))
```

Bhasan Char is in Noakhali district (Chattogram division).
Cox's Bazar district suspected cases to cutoff: `r format(cum_table$suspected[cum_table$place == "Cox's Bazar district"], big.mark = ",")`.
Whether camp cases are in the DGHS Cox's Bazar count is unknown (Q12).

# Severity

All DGHS cases are hospital admissions, so fatality ratios are hospital fatality ratios (HFR).

```{r}
#| label: inputs-severity

weekly_sev <- div_daily |>
  complete_weeks() |>
  group_by(division, week_end) |>
  summarise(across(c(suspected, confirmed, suspected_deaths,
                     confirmed_deaths, admitted, discharged), sum),
            .groups = "drop") |>
  mutate(deaths_A = suspected_deaths,
         deaths_B = suspected_deaths + confirmed_deaths)
```

```{r}
#| label: fig-admission-ratio
#| fig-cap: "Admissions per suspected case, by division and week, Bangladesh, weeks ending 14 April 2026 to cutoff. Source: DGHS measles monitoring platform, report date."
#| fig-alt: "Line charts, one panel per division, of weekly admissions divided by suspected cases."

p <- weekly_sev |>
  mutate(ratio = admitted / suspected) |>
  ggplot(aes(week_end, ratio)) +
  geom_hline(yintercept = 1, colour = col_reference, linetype = "dotted") +
  geom_line(colour = col_primary) +
  facet_wrap(~division, ncol = 4) +
  scale_x_date(date_labels = "%b") +
  labs(x = "Week ending", y = "Admitted / suspected")
ggplotly(p)
```

```{r}
#| label: fig-hfr-resolved
#| fig-cap: "HFR among resolved admissions, deaths / (deaths + discharges), by division and week, under reading A (suspected deaths include confirmed, black) and B (separate, red), Bangladesh, weeks ending 14 April 2026 to cutoff. Valid only if deaths are hospital deaths (Q7). Source: DGHS measles monitoring platform, report date."
#| fig-alt: "Line charts, one panel per division, of weekly hospital fatality among resolved admissions under two readings of the death counts."

hfr_resolved <- weekly_sev |>
  mutate(A = deaths_A / (deaths_A + discharged),
         B = deaths_B / (deaths_B + discharged)) |>
  pivot_longer(c(A, B), names_to = "reading", values_to = "hfr")
stopifnot(all(between(hfr_resolved$hfr, 0, 1), na.rm = TRUE))

p <- ggplot(hfr_resolved, aes(week_end, hfr, colour = reading)) +
  geom_line() +
  facet_wrap(~division, ncol = 4) +
  scale_colour_manual(values = c(A = col_primary, B = col_highlight)) +
  scale_x_date(date_labels = "%b") +
  labs(x = "Week ending", y = "Deaths / (deaths + discharges)",
       colour = "Reading")
ggplotly(p)
```

```{r}
#| label: fig-deaths
#| fig-cap: "Suspected and confirmed deaths per week, national, Bangladesh, weeks ending 14 April 2026 to cutoff. Source: DGHS measles monitoring platform, report date."
#| fig-alt: "Line chart of weekly suspected deaths and confirmed deaths."

deaths_weekly <- nat_daily |>
  complete_weeks() |>
  group_by(week_end) |>
  summarise(`Suspected deaths` = sum(suspected_deaths),
            `Confirmed deaths` = sum(confirmed_deaths)) |>
  pivot_longer(-week_end, names_to = "measure", values_to = "deaths")
p <- ggplot(deaths_weekly, aes(week_end, deaths, colour = measure)) +
  geom_line() +
  geom_point(size = 1) +
  scale_colour_manual(values = c(`Suspected deaths` = col_primary,
                                 `Confirmed deaths` = col_highlight)) +
  scale_x_date(date_labels = "%b") +
  labs(x = "Week ending", y = "Deaths per week", colour = NULL)
ggplotly(p)
```

```{r}
#| label: tbl-hfr
#| tbl-cap: "Naive HFR, cumulative deaths / cumulative cases, by division, under readings A and B, Bangladesh, 15 March 2026 to cutoff. Source: DGHS measles monitoring platform."

hfr_naive <- cum_table |>
  filter(place %in% c(division_order, "National")) |>
  transmute(
    place, suspected, confirmed, suspected_deaths, confirmed_deaths,
    hfr_A_pct = 100 * suspected_deaths / suspected,
    hfr_B_pct = 100 * (suspected_deaths + confirmed_deaths) /
      (suspected + confirmed)
  )
stopifnot(all(hfr_naive$suspected_deaths <= hfr_naive$suspected))
kable(hfr_naive, digits = 2, format.args = list(big.mark = ","))
```

```{r}
#| label: fig-wasting
#| fig-cap: "Wasting among children under five (BDHS 2022) and naive HFR (reading A), by division, Bangladesh. Shown side by side, not fitted. Sources: DHS Program subnational indicators; DGHS measles monitoring platform."
#| fig-alt: "Two dot plots side by side, divisions on the vertical axis: percentage of children wasted, and hospital fatality ratio."
#| fig-height: 4

dhs_path <- here("data", "covariates", "dhs_division.csv")
if (file.exists(dhs_path)) {
  wasting <- read.csv(dhs_path) |>
    filter(indicator_short == "wasting") |>
    transmute(place = division, value = value,
              panel = paste0("Wasting, under five (%, ", survey, ")"))
  hfr_dots <- hfr_naive |>
    filter(place != "National") |>
    transmute(place = as.character(place), value = hfr_A_pct,
              panel = "Naive HFR, reading A (%)")
  bind_rows(wasting, hfr_dots) |>
    mutate(place = factor(place, levels = rev(division_order))) |>
    ggplot(aes(value, place)) +
    geom_point(colour = col_primary) +
    facet_wrap(~panel, scales = "free_x") +
    labs(x = NULL, y = NULL)
}
```

## Delay-adjusted HFR

Not run.
Waits on agreement of the delay from report to death report (`docs/describe-spec.md`, Severity).

```{r}
#| label: tbl-hfr-delay
#| eval: false

# cfr::cfr_static() and cfr::cfr_rolling() by division, over a grid of
# gamma delays (mean 5, 10, 15 days; CV 0.5), readings A and B.
```
