---
title: "Measles outbreak, Bangladesh 2026: descriptive epidemiology"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-tools: true
    fig-width: 9
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup

library(here)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(ggridges)
library(plotly)
library(sf)
library(slider)
library(knitr)

# visualise-data palette
col_primary <- "#1a1a1a"
col_highlight <- "#c0392b"
col_reference <- "#7f7f7f"
# Okabe-Ito, one colour per division
col_divisions <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442",
                   "#0072B2", "#D55E00", "#CC79A7", "#999999")
theme_set(
  theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank(), legend.position = "bottom")
)
sf_use_s2(FALSE)

# Utility functions
# MMWR epidemiological weeks, Sunday to Saturday, labelled by start date
week_start_of <- function(date) floor_date(date, "week", week_start = 7)
week_label <- function(week) {
  paste0("week ", epiweek(week), " (", fmt_date(week), " to ",
         fmt_date(week + 6), ")")
}
# Sum daily counts by complete epiweek, grouped by the columns in ...
weekly <- function(d, ...) {
  d |>
    mutate(week = week_start_of(date)) |>
    filter(between(week, first_week, last_week)) |>
    group_by(..., week) |>
    summarise(across(c(suspected, confirmed, total, serum_sent), sum),
              .groups = "drop")
}
fmt <- function(x) format(round(x), big.mark = ",")
fmt_date <- function(d, year = FALSE) {
  paste(day(d), format(d, if (year) "%B %Y" else "%B"))
}
```

```{r}
#| label: inputs

# DGHS dashboard series by report date
cases <- read.csv(here("data", "dghs-cases.csv"), na.strings = c("", "NA")) |>
  mutate(date = ymd(date))
cutoff <- max(cases$date)
first_report <- ymd("2026-03-15")
first_daily <- ymd("2026-04-08")
# First and last complete epiweeks of daily data
first_week <- week_start_of(first_daily - 1) + 7
last_week <- week_start_of(cutoff + 1) - 7

# DGHS district spellings that differ from the OCHA boundaries
district_recode <- c(
  Jhalakathi = "Jhalokati",
  Netrokona = "Netrakona",
  Chapainawabganj = "Chapainababganj"
)
cases <- cases |>
  mutate(district = recode(district, !!!district_recode))

# Boundaries and population (2022 census projection)
adm1 <- st_read(here("data", "covariates", "bgd_adm1.geojson"), quiet = TRUE)
adm2 <- st_read(here("data", "covariates", "bgd_adm2.geojson"), quiet = TRUE)
lookup <- read.csv(here("data", "covariates", "admin_lookup.csv"))

# Divisions west to east by centroid, and districts the same way within each
centroid_lon <- function(g) {
  suppressWarnings(st_coordinates(st_centroid(g))[, 1])
}
division_order <- adm1 |>
  mutate(lon = centroid_lon(geometry)) |>
  arrange(lon) |>
  pull(adm1_name)
district_order <- adm2 |>
  mutate(
    lon = centroid_lon(geometry),
    division = factor(adm1_name, levels = division_order)
  ) |>
  arrange(division, lon) |>
  pull(adm2_name)

pop1 <- read.csv(here("data", "covariates", "population_adm1.csv")) |>
  select(adm1_pcode, pop = t_tl, pop_u5 = t_00_04) |>
  left_join(distinct(lookup, adm1_pcode, division = adm1_name),
            by = "adm1_pcode")
pop2 <- read.csv(here("data", "covariates", "population_adm2.csv")) |>
  select(adm2_pcode, pop = t_tl, pop_u5 = t_00_04) |>
  left_join(select(lookup, adm2_pcode, district = adm2_name,
                   division = adm1_name),
            by = "adm2_pcode")
stopifnot(
  setequal(pop1$division, unique(na.omit(cases$division))),
  setequal(pop2$district, unique(na.omit(cases$district)))
)

# Wide daily series: one row per date and place, one column per measure.
# Total = suspected + confirmed, treating the two as separate counts
widen <- function(d) {
  d |>
    select(-fetched_at, -flag) |>
    pivot_wider(names_from = measure, values_from = value) |>
    mutate(total = suspected + confirmed,
           total_deaths = suspected_deaths + confirmed_deaths)
}
nat_daily <- cases |>
  filter(level == "national") |>
  widen()
div_daily <- cases |>
  filter(level == "division") |>
  widen() |>
  mutate(division = factor(division, levels = division_order))
dist_daily <- cases |>
  filter(level == "district") |>
  widen() |>
  mutate(district = factor(district, levels = district_order),
         division = factor(division, levels = division_order))
# 15 March to 7 April, loaded onto 7 April as one total
lump <- nat_daily |> filter(date == first_daily - 1)
lump_days <- as.numeric(first_daily - first_report)

# Events marked on time series
events <- read.csv(here("assets", "events.csv")) |>
  mutate(date = ymd(date))
events_marked <- events |>
  filter(!grepl("backfill", event))

# WHO provisional monthly measles and rubella data
who <- read.csv(here("data", "who-monthly.csv")) |>
  mutate(month = make_date(year, month, 1)) |>
  filter(year >= 2023)
```

::: callout-note
- Data to `r fmt_date(cutoff, year = TRUE)`, from the DGHS measles monitoring platform ([measles.dghs.gov.bd](https://measles.dghs.gov.bd)).
  Many thanks to DGHS for making these data public.
- A case is a hospital admission with suspected or confirmed measles.
  In this analysis, total cases = suspected + confirmed.
:::

::: {.callout-note collapse="true"}
# Data and definitions

## Data reporting process

- Facilities report measles admissions daily, collated by district civil surgeon offices and submitted as daily counts to the DGHS platform.
- Daily surveillance reporting start on 8 April.
  Cases from 15 March to 7 April were accumulated onto 7 April as one total (`r fmt(lump$total)` cases: `r fmt(lump$suspected)` suspected, `r fmt(lump$confirmed)` confirmed).
- Dates are report dates, and place is the district of the reporting facility.
- Weeks are MMWR epidemiological weeks, Sunday to Saturday.
  Weekly figures show complete weeks only.
- Population is the 2022 census projection.

### Source data variables

```{r}
#| label: tbl-measures
#| tbl-cap: "DGHS dashboard indicators, reported as 24-hour counts by report date"

tibble::tribble(
  ~measure, ~dashboard_label, ~notes,
  "suspected", "suspected24h", "Hospital admissions with suspected measles; whether confirmed are included is not stated",
  "confirmed", "confirmed24h", "Laboratory method not stated",
  "total", "(derived)", "suspected + confirmed; assumes the two are separate",
  "suspected_deaths", "suspectedDeath24h", "Whether confirmed deaths are included, and place of death, not stated",
  "confirmed_deaths", "confirmedDeath24h", "As above",
  "total_deaths", "(derived)", "suspected_deaths + confirmed_deaths; assumes the two are separate",
  "admitted", "admitted24h", "Relation to suspected not stated",
  "discharged", "discharged24h", "\"Recovered\" in the dashboard's Excel export",
  "serum_sent", "serumSent24h", "Serum samples sent to the laboratory; not in the Excel export"
) |>
  kable(col.names = c("Measure", "Dashboard label", "Notes"))
```

### WHO standard measles case definitions

```{r}
#| label: tbl-who-definitions
#| tbl-cap: "WHO measles surveillance case definitions, used in WHO monthly data (verbatim)"

read.csv(here("assets", "case-definitions.csv")) |>
  select(classification, definition) |>
  kable(col.names = c("Classification", "Definition"))
```
:::

# Situation update

```{r}
#| label: situation

cum <- nat_daily |>
  summarise(across(c(suspected, confirmed, total, suspected_deaths,
                     confirmed_deaths, total_deaths), sum))
nat_weekly <- weekly(nat_daily) |>
  mutate(per_100k = 1e5 * total / sum(pop1$pop))
this_week <- nat_weekly |> filter(week == last_week)
prev_week <- nat_weekly |> filter(week == last_week - 7)
change_pct <- 100 * (this_week$total / prev_week$total - 1)
div_last <- weekly(div_daily, division) |>
  filter(week == last_week) |>
  left_join(pop1, by = "division") |>
  mutate(per_100k = 1e5 * total / pop) |>
  arrange(desc(per_100k))
dist_last <- weekly(dist_daily, district) |>
  filter(week == last_week)
```

- Cumulative, 15 March to `r fmt_date(cutoff)`: `r fmt(cum$total)` cases (`r fmt(cum$suspected)` suspected, `r fmt(cum$confirmed)` confirmed) and `r fmt(cum$total_deaths)` deaths (`r fmt(cum$suspected_deaths)` suspected, `r fmt(cum$confirmed_deaths)` confirmed).
- Latest complete `r week_label(last_week)`: `r fmt(this_week$total)` cases, `r sprintf("%+.0f%%", change_pct)` on the week before, `r sprintf("%.1f", this_week$per_100k)` per 100,000 population (@fig-epicurve-national).
- `r sum(dist_last$total > 0)` of 64 districts reported a case in the latest week.
  Highest division incidence: `r div_last$division[1]`, `r sprintf("%.1f", div_last$per_100k[1])` per 100,000.

## Epidemic curve

The outbreak was alerted to public health surveillance on 15 March, with daily counts reported from 8 April.
The outbreak likely began earlier.

```{r}
#| label: fig-epicurve-national
#| fig-cap: >-
#|   Measles hospital admissions per epidemiological week, confirmed and suspected, Bangladesh, 15 March 2026 to the latest complete week.
#|   Shaded block: 15 March to 7 April, reported as one total and shown at its weekly average; 8 to 11 April, a partial week, is not shown.
#|   Dotted lines: campaign start in high-risk upazilas (5 April); nationwide campaign (20 April); camp campaign and platform launch (26 April).
#|   Source: DGHS measles monitoring platform, report date.
#| fig-alt: "Weekly stacked bar chart of national confirmed and suspected measles admissions, with a shaded block for the 15 March to 7 April total and dotted lines at campaign dates."

type_levels <- c("Suspected", "Confirmed")
type_colours <- c(Suspected = col_reference, Confirmed = col_highlight)
nat_weekly_long <- nat_weekly |>
  mutate(text = paste0(week_label(week), ": ", fmt(total), " cases (",
                       fmt(suspected), " suspected, ", fmt(confirmed),
                       " confirmed)")) |>
  pivot_longer(c(suspected, confirmed), names_to = "type",
               values_to = "cases") |>
  mutate(type = factor(tools::toTitleCase(type), levels = type_levels))

# Lumped total as a weekly average, confirmed below suspected
lump_rate <- 7 / lump_days
lump_rect <- tibble(
  type = factor(type_levels, levels = type_levels),
  ymin = c(lump$confirmed, 0) * lump_rate,
  ymax = c(lump$total, lump$confirmed) * lump_rate,
  text = paste0("15 March to 7 April, reported as one total: ",
                fmt(lump$total), " cases (", fmt(lump$suspected),
                " suspected, ", fmt(lump$confirmed), " confirmed)")
)

p <- ggplot() +
  geom_rect(data = lump_rect,
            aes(xmin = first_report, xmax = first_daily, ymin = ymin,
                ymax = ymax, fill = type, text = text),
            alpha = 0.35, colour = col_primary, linetype = "dashed",
            linewidth = 0.3, show.legend = FALSE) +
  geom_col(data = nat_weekly_long,
           aes(week + 3.5, cases, fill = type, text = text), width = 7) +
  geom_vline(data = filter(events_marked, date >= first_report),
             aes(xintercept = date), colour = col_primary,
             linetype = "dotted") +
  annotate("text", x = first_report, y = max(lump_rect$ymax),
           label = "Reported as\none total", hjust = 0, vjust = -0.3,
           size = 3, colour = col_reference) +
  scale_fill_manual(values = type_colours) +
  scale_x_date(date_labels = "%b") +
  labs(x = "Epiweek", y = "Cases per week", fill = NULL)
ggplotly(p, tooltip = "text")
```

Epidemiological intervals (i.e. the incubation period, the time from symptom onset to medical admission, and the time from admission to data report): see [Appendix A](#sec-timespans).

::: {.callout-tip collapse="true"}
# Last 7 days' data extract

```{r}
#| label: latest-daily

nat_daily |>
  select(date, suspected, confirmed, total, suspected_deaths,
         confirmed_deaths, admitted, discharged, serum_sent) |>
  tail(7) |>
  kable(caption = "Daily DGHS measles dashboard data, national",
        format.args = list(big.mark = ","))
```
:::

--------------------------------------------------------------------------------

# Spatial spread {#sec-spatial}

```{r}
#| label: inputs-where

district_cum <- dist_daily |>
  group_by(district) |>
  summarise(total = sum(total), .groups = "drop") |>
  left_join(pop2, by = "district") |>
  mutate(per_100k = 1e5 * total / pop)
stopifnot(nrow(district_cum) == 64, !anyNA(district_cum$pop))

map_districts <- adm2 |>
  left_join(district_cum, by = c("adm2_name" = "district"))
breaks_100k <- c(0, 25, 50, 100, 200, 400, 800)
stopifnot(!anyNA(map_districts$per_100k),
          max(map_districts$per_100k) <= max(breaks_100k))
```

```{r}
#| label: fig-map-cumulative
#| fig-cap: >-
#|   Cumulative measles cases (suspected + confirmed) per 100,000 population, by district of reporting facility, 15 March 2026 to cutoff.
#|   Sources: DGHS measles monitoring platform; 2022 census projection.
#| fig-alt: "Choropleth map of Bangladesh districts shaded by cumulative measles cases per 100,000, with division outlines."
#| fig-height: 8

ggplot() +
  geom_sf(data = map_districts, aes(fill = per_100k), colour = "white",
          linewidth = 0.1) +
  geom_sf(data = adm1, fill = NA, colour = col_primary, linewidth = 0.4) +
  scale_fill_viridis_b(breaks = breaks_100k, limits = range(breaks_100k),
                       option = "magma", direction = -1,
                       show.limits = TRUE) +
  labs(fill = "Cases\nper 100,000") +
  theme_void() +
  theme(legend.position = "right")
```

```{r}
#| label: fig-ridge-district
#| fig-cap: >-
#|   Weekly measles cases (suspected + confirmed) per 100,000 population, by district of reporting facility, from 12 April 2026 to the latest complete week.
#|   Districts ordered by week of peak incidence.
#|   Sources: DGHS measles monitoring platform, report date; 2022 census projection.
#| fig-alt: "Ridgeline plot of weekly measles cases per 100,000 for 64 districts, coloured by division, ordered by week of peak incidence."
#| fig-height: 14

# Incidence 15 March to 7 April, the earliest district data
pre_platform <- dist_daily |>
  filter(date == first_daily - 1) |>
  left_join(select(pop2, district, pop), by = "district") |>
  transmute(district = as.character(district),
            pre_100k = 1e5 * total / pop)
dist_weekly <- weekly(dist_daily, division, district) |>
  left_join(select(pop2, district, pop), by = "district") |>
  mutate(per_100k = 1e5 * total / pop)
# Factor levels run bottom to top: latest peak first, so the earliest
# peak is the top row
ridge_order <- dist_weekly |>
  group_by(district = as.character(district)) |>
  summarise(peak_week = week[which.max(per_100k)], .groups = "drop") |>
  left_join(pre_platform, by = "district") |>
  arrange(desc(peak_week), pre_100k) |>
  pull(district)
# One row of height on the y axis, in cases per 100,000 a week
row_100k <- signif(max(dist_weekly$per_100k) / 3, 1)

dist_weekly |>
  ggplot(aes(week + 3.5, factor(district, levels = ridge_order),
             height = per_100k, fill = division)) +
  geom_ridgeline(scale = 1 / row_100k, alpha = 0.7, colour = col_primary,
                 linewidth = 0.2) +
  scale_fill_manual(values = col_divisions) +
  scale_x_date(date_labels = "%b", expand = c(0, 0)) +
  # Room above the top ridges
  scale_y_discrete(expand = expansion(add = c(0.5, 2.5))) +
  labs(x = "Week (Sunday to Saturday)", y = NULL, fill = "Division",
       subtitle = paste("Height of one row =", row_100k,
                        "cases per 100,000 a week")) +
  theme(axis.text.y = element_text(size = 7),
        panel.grid.major.y = element_blank())
```

## Cox's Bazar and refugee camps {#sec-camps}

Refugee camps report separately from the DGHS platform, where camp health facilities report syndromic surveillance directly to the Rohingya Response early warning system (EWARS).

```{r}
#| label: inputs-camps

# EWARS weekly suspected measles/rubella in the camps, digitised from the
# W36 2026 bulletin chart; confirmed as a series (data/camps-review.csv)
ewars <- read.csv(here("data", "camps-ewars-weekly.csv")) |>
  filter(confirmed_by != "") |>
  transmute(week = ymd(week_start), ewars = value)
# DGHS admissions reported by facilities in Cox's Bazar district
coxs_weekly <- dist_daily |>
  filter(district == "Cox's Bazar") |>
  weekly() |>
  select(week, dghs = total)
camps_weekly <- full_join(ewars, coxs_weekly, by = "week") |>
  arrange(week)

# Camp case and death figures from situation reports and bulletins,
# confirmed against the source page
camps_figures <- read.csv(here("data", "camps-measles.csv"),
                          na.strings = "") |>
  filter(!is.na(confirmed_by),
         indicator %in% c("camp_suspected_cases", "camp_confirmed_cases",
                          "camp_epilinked_cases", "camp_total_cases",
                          "camp_deaths", "camp_admissions")) |>
  mutate(date = ymd(coalesce(as_of_date, published_date)))
```

```{r}
#| label: fig-epicurve-camps
#| fig-cap: >-
#|   Suspected measles cases per epidemiological week in the refugee camps (bars), and measles admissions (suspected + confirmed) reported by facilities in Cox's Bazar district (line), 22 December 2024 to 12 September 2026.
#|   Dotted line: camp MR campaign (26 April).
#|   Sources: Rohingya Response EWARS bulletin W36 2026 (Figure 4, digitised); DGHS measles monitoring platform, report date.
#| fig-alt: "Weekly bar chart of suspected measles cases in the refugee camps from late 2024, with a line of weekly measles admissions in Cox's Bazar district from April 2026 and a dotted line at the camp campaign."

p <- camps_weekly |>
  mutate(text = paste0(week_label(week), ": ",
                       if_else(is.na(ewars), "", paste0(fmt(ewars), " camps")),
                       if_else(is.na(dghs), "",
                               paste0("; ", fmt(dghs), " Cox's Bazar (DGHS)")))) |>
  ggplot(aes(week + 3.5)) +
  geom_col(aes(y = ewars, fill = "Camps (EWARS)", text = text), width = 7,
           na.rm = TRUE) +
  geom_line(aes(y = dghs, colour = "Cox's Bazar district (DGHS)"),
            na.rm = TRUE) +
  geom_point(aes(y = dghs, colour = "Cox's Bazar district (DGHS)",
                 text = text), size = 1, na.rm = TRUE) +
  geom_vline(data = filter(events, scope == "camps"),
             aes(xintercept = date), colour = col_primary,
             linetype = "dotted") +
  scale_fill_manual(values = c("Camps (EWARS)" = col_reference)) +
  scale_colour_manual(values = c("Cox's Bazar district (DGHS)" =
                                   col_highlight)) +
  scale_x_date(date_labels = "%b %Y") +
  labs(x = "Week (Sunday to Saturday)", y = "Cases per week",
       fill = NULL, colour = NULL)
ggplotly(p, tooltip = "text")
```

::: {.callout-tip collapse="true"}
# Camp case and death figures in situation reports

```{r}
#| label: tbl-camps-figures
#| tbl-cap: >-
#|   Indicators of measles cases and deaths in Cox's Bazar refugee camps.
#|   Note that geography, period and denominator are not reported consistently and differ between documents, so are presented as stated.
#|   Sources: Health Sector Cox's Bazar bulletins; UN RCO, UNICEF and WHO SEARO reports.

camps_figures |>
  arrange(indicator, date) |>
  transmute(
    measure = sub("^camp_", "", indicator) |> gsub(pattern = "_", replacement = " "),
    value = fmt(value),
    date = if_else(is.na(date), "", format(date, "%e %b")),
    geography, period = coalesce(period_text, period, ""),
    source = doc_id, note = coalesce(gsub("\n", " ", review_note), "")
  ) |>
  kable(col.names = c("Measure", "Value", "As of or published", "Geography",
                      "Period", "Document", "Reviewer note"))
```
:::

--------------------------------------------------------------------------------

# Lab testing {#sec-testing}

The DGHS platform records volume of serum samples.
This isn't proper test positivity, as samples sent aren't directly linked to confirmed cases.

```{r}
#| label: tbl-testing
#| tbl-cap: >-
#|   Measles cases, serum samples sent and share confirmed, by division, Bangladesh, 1 May 2026 to cutoff.
#|   Source: DGHS measles monitoring platform, report date.

serum_start <- ymd("2026-05-01")
sum_testing <- function(d) {
  d |>
    filter(date >= serum_start) |>
    summarise(across(c(suspected, confirmed, total, serum_sent), sum))
}
bind_rows(
  div_daily |>
    group_by(place = as.character(division)) |>
    sum_testing(),
  nat_daily |>
    sum_testing() |>
    mutate(place = "National")
) |>
  mutate(serum_per_case = serum_sent / total,
         share_confirmed = confirmed / total) |>
  kable(digits = 2, format.args = list(big.mark = ","),
        col.names = c("Place", "Suspected", "Confirmed", "Total",
                      "Serum sent", "Serum sent per case",
                      "Share confirmed"))
```

There is separate data on lab testing from WHO monthly surveillance, which also records 'discarded' negative cases.
This is an upper bound as cases can also be discarded without a test (@tbl-who-definitions).

```{r}
#| label: tbl-who-positivity
#| tbl-cap: >-
#|   Measles cases by final classification, share tested and laboratory positivity, by month.
#|   Share tested = (laboratory-confirmed + discarded) / suspected.
#|   Laboratory positivity = laboratory-confirmed / (laboratory-confirmed + discarded).
#|   Source: WHO provisional monthly measles and rubella data.

who |>
  filter(year == 2026) |>
  transmute(
    month = format(month, "%B"),
    suspected = measles_suspect,
    lab_confirmed = measles_lab_confirmed,
    epi_linked = measles_epi_linked,
    clinical = measles_clinical,
    discarded,
    share_tested = (measles_lab_confirmed + discarded) / measles_suspect,
    positivity = measles_lab_confirmed / (measles_lab_confirmed + discarded)
  ) |>
  kable(digits = 2, format.args = list(big.mark = ","),
        col.names = c("Month", "Suspected", "Lab-confirmed", "Epi-linked",
                      "Clinically compatible", "Discarded", "Share tested",
                      "Lab positivity"))
```

# Vaccination campaigns

```{r}
#| label: tbl-campaigns
#| tbl-cap: "Measles-rubella vaccination campaign rounds, Bangladesh, 2026"

events |>
  filter(grepl("campaign", event, ignore.case = TRUE)) |>
  transmute(date = fmt_date(date, year = TRUE), event, scope, source) |>
  kable(col.names = c("Date", "Event", "Scope", "Source"))
```

Dates of the emergency and mop up round TBC.

--------------------------------------------------------------------------------

# Appendix A: Epidemiologic time intervals {#sec-timespans}

```{r}
#| label: tbl-timespans
#| tbl-cap: "Epidemiological time intervals and potential data sources"

tibble::tribble(
  ~interval, ~public_data, ~line_list_variable, ~other_source,
  "Generation or serial interval", "No", "No (no transmission pairs)", "Literature",
  "Exposure to onset (incubation)", "No", "No", "Literature",
  "Fever to rash", "No", "Yes", "",
  "Rash to notification", "No", "Yes", "",
  "Notification to investigation", "No", "Yes", "",
  "Rash to specimen; specimen to result", "No", "Yes", "",
  "Onset to admission", "No", "Yes", "",
  "Admission to report date", "No; indirect signs only (days with missing reports)", "No", "",
  "Suspected to confirmed", "No (share tested varies)", "Yes", "",
  "Admission to exit (discharge or death)", "Partly: deconvolution of discharges and deaths against admissions", "Yes", "",
  "Onset to death", "No", "Free-text Comment only", "Unknown",
  "Campaign dose to rash", "No", "DateLastMCV", "",
  "Undetected period", "No", "Onset dates before April", "Molecular clock on sequences"
) |>
  kable(col.names = c("Interval", "Public data", "Line list variable",
                      "Other source"))
```

## Day of week reporting

```{r}
#| label: tbl-weekday
#| tbl-cap: >-
#|   Daily measles cases (suspected + confirmed) relative to the centred 7-day mean, by weekday and division: median (interquartile range), Bangladesh, 8 April 2026 to cutoff.
#|   Source: DGHS measles monitoring platform, report date.

weekday <- div_daily |>
  filter(date >= first_daily) |>
  arrange(division, date) |>
  group_by(division) |>
  mutate(centred = slide_dbl(total, mean, .before = 3, .after = 3,
                             .complete = TRUE),
         rel = total / centred) |>
  ungroup() |>
  filter(is.finite(rel)) |>
  mutate(weekday = wday(date, label = TRUE, week_start = 6)) |>
  group_by(division, weekday) |>
  summarise(cell = sprintf("%.2f (%.2f to %.2f)", median(rel),
                           quantile(rel, 0.25), quantile(rel, 0.75)),
            .groups = "drop") |>
  pivot_wider(names_from = weekday, values_from = cell)
kable(weekday)
```

# Appendix B: Data validation {#sec-validation}

## Comparison with WHO monthly surveillance

WHO monthly surveillance is a separate stream, built from EPI case investigations.
Its case classification appears to change from June, when epidemiological linking stops and nearly every suspected case is tested (@tbl-who-positivity).

```{r}
#| label: fig-timeline
#| fig-cap: >-
#|   Measles cases per month: WHO surveillance suspected and total cases (bars), and DGHS hospital admissions, suspected + confirmed (points), Bangladesh, January 2026 to cutoff.
#|   DGHS April is 8 to 30 April; 15 March to 7 April is reported as one total; the current month is shown when complete.
#|   Sources: WHO provisional monthly measles and rubella data; DGHS measles monitoring platform.
#| fig-alt: "Monthly bar chart of WHO suspected and total measles cases with DGHS monthly hospital admissions as points."

who_2026 <- who |>
  filter(year(month) == 2026) |>
  select(month, `WHO suspected` = measles_suspect,
         `WHO total (lab + epi-linked + clinical)` = measles_total) |>
  pivot_longer(-month, names_to = "series", values_to = "cases")
dghs_monthly <- nat_daily |>
  filter(date >= first_daily) |>
  group_by(month = floor_date(date, "month")) |>
  summarise(cases = sum(total), days = n(), .groups = "drop") |>
  # Current month dropped while incomplete; April starts on 8 April
  filter(month < floor_date(cutoff + 1, "month")) |>
  mutate(series = "DGHS total (hospital)")

p <- ggplot() +
  geom_col(data = who_2026,
           aes(month, cases, fill = series),
           position = position_dodge(width = 25), width = 24) +
  geom_point(data = dghs_monthly,
             aes(month, cases, colour = series, text = paste(days, "days")),
             size = 2.5) +
  geom_vline(xintercept = ymd("2026-06-01"), linetype = "dotted",
             colour = col_reference) +
  annotate("text", x = ymd("2026-06-01"), y = max(dghs_monthly$cases),
           label = "WHO classification change", hjust = -0.05,
           size = 3, colour = col_reference) +
  scale_fill_manual(values = c(
    `WHO suspected` = "#d4d4d4",
    `WHO total (lab + epi-linked + clinical)` = col_reference)) +
  scale_colour_manual(values = c(`DGHS total (hospital)` = col_highlight)) +
  scale_x_date(date_labels = "%b") +
  labs(x = NULL, y = "Cases per month", fill = NULL, colour = NULL)
ggplotly(p, tooltip = c("x", "y", "text"))
```

## Data quality

Checks that failed when the dashboard series was tidied, and dates flagged during tidying.

```{r}
#| label: tbl-data-quality
#| tbl-cap: "Failed checks and flagged dates, DGHS dashboard series, Bangladesh, 2026"

checks <- read.csv(here("data", "dghs-cases-checks.csv"))
failed <- checks |>
  filter(!pass) |>
  transmute(item = check, date, division, measure,
            detail = paste("expected", expected, "observed", observed))
flagged <- cases |>
  filter(!is.na(flag), level == "national", measure == "suspected") |>
  transmute(item = flag, date = as.character(date), division = NA,
            measure = NA, detail = paste("national suspected", value))
bind_rows(failed, flagged) |> kable()
```

Daily dashboard counts compared with the same counts in DGHS press releases.

```{r}
#| label: tbl-source-agreement
#| tbl-cap: "Agreement of dashboard and press release counts by month, Bangladesh, 2026"

read.csv(here("sitrep", "data", "dghs-source-compare.csv")) |>
  mutate(month = format(ymd(date), "%Y-%m")) |>
  filter(!is.na(diff)) |>
  group_by(month) |>
  summarise(comparisons = n(), differ = sum(diff != 0),
            net_difference = sum(diff)) |>
  kable(col.names = c("Month", "Comparisons", "Differ", "Net difference"))
```
