---
title: "Data description"
format:
  html:
    embed-resources: true
    code-fold: true
---

- Scope: Familiarity and basic descriptive analysis of public data.
- Authored: Kath Sherratt 1 Oct 2026
- Context: Exploring public data downloads relevant to the 2026 measles outbreak.

```{r}
#| label: set-up

library(here)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
theme_set(theme_bw() + theme(legend.position = "bottom"))
```

## Current outbreak

### Daily DGHS data

Outbreak case data from the DGHS.
The DGHS started providing daily sitreps on 2 April 2026, and then a dashboard with daily numbers from 26 April.

```{r}
#| label: dghs-dashboard
#| fig-cap: "DGHS daily reporting, 2026; black = total admissions, blue = lab-confirmed admissions, grey = suspected cases"

dghs <- read.csv(here("data", "dghs-cases.csv"))
dghs <- dghs |> mutate(date = ymd(date))
dghs_national <- filter(dghs, level == "national") |>
    select(date, measure, value)
tail(dghs_national)

dghs_national |>
    filter(measure %in% c("confirmed", "admitted", "suspected")) |>
    filter(date >= "2026-04-26") |>
    pivot_wider(names_from = measure, values_from = value) |>
    mutate(total = confirmed + admitted) |>
    ggplot(aes(x = date)) +
    geom_point(aes(y = total)) +
    geom_line(aes(y = total)) +
    geom_point(aes(y = confirmed), colour = "blue", alpha = 0.5) +
    geom_point(aes(y = suspected), colour = "grey", alpha = 0.5) +
    labs(y = NULL, x = NULL)
```

We know the outbreak started much earlier than this.
The daily sitreps start by reporting a cumulative outbreak size from 15 March to 2 April that totalled 7467 admissions in that period:

```{r}
#| label: pre-dashboard-cumulative

dghs_national |> filter(date == ymd("2026-04-07")) |> select(-date)
```

Looking at all this cumulatively in the population:

```{r}
# TODO cumulative plot per 100k population
```

### Hospital admissions and deaths

The DGHS dashboard counts come from medical facilities reporting hospital admissions.
This includes suspected, clinically confirmed, and lab-confirmed measles admissions; confirmed deaths, and suspected deaths.

```{r}
#| label: dghs-case-death

case_death <- dghs_national |>
    filter(measure %in% c("confirmed", "confirmed_deaths")) |>
    filter(date >= "2026-04-26") |>
    pivot_wider(names_from = measure, values_from = value) |>
    mutate(death_case = confirmed_deaths / confirmed)
tail(case_death)

case_death |> 
    ggplot(aes(x = date)) +
    geom_point(aes(y = confirmed_deaths)) +
    geom_line(aes(y = confirmed)) +
    labs(y = NULL, x = NULL)
```

### Lab testing

We can roughly guess at test positivity by looking at share of confirmed cases to serum-samples sent for testing.^[I'm not clear whether confirmed-deaths are the outcome of confirmed-cases or if these were tested separately.
I am excluding these to avoid risk of double-counting.]

```{r}
#| label: dghs-lab-confirmation

test_pos <- dghs_national |>
    filter(measure %in% c("confirmed", "serum_sent")) |>
    filter(date >= "2026-04-26") |>
    pivot_wider(names_from = measure, values_from = value) |>
    mutate(test_pos = confirmed / serum_sent)
tail(test_pos)

#TODO check and note the early spike (backfill?)

test_pos |> 
    ggplot(aes(x = date)) +
    geom_point(aes(y = test_pos)) +
    geom_line(aes(y = test_pos)) +
    geom_smooth(aes(y = test_pos), method = "gam", formula = y ~ s(x, bs = "cs"), 
    alpha = 0.5, col = "grey") +
    labs(y = NULL, x = NULL)
```

## Spatial spread

--------------------------------------------------------------------------------

## Public health response

### Pre-outbreak vaccination coverage

### Responsive vaccination campaign

--------------------------------------------------------------------------------

## Context

### Determinants of severe disease

#### Malnutrition prevalence, regional

```{r}
#| label: malnutrition-map
```

### Historic trends

#### Measles case reporting

WHO provides a monthly timeseries of case reports since 2012.
The "total" case count adds epi-linked, clinically diagnosed, and lab-confirmed cases.
We also have a count of suspected cases.

```{r}
#| label: who-historic

who <- read.csv(here("data", "who-monthly.csv"))
who_long <- who |> 
    select("year", "month", "measles_total", "measles_suspect") |>
    mutate(date = make_date(year, month, 1))
tail(who_long)
```

For a baseline comparison, we plot this over the past few years up to October 2025:

```{r}
#| label: who-historic-plot
#| fig-cap: "WHO monthly measles case reporting, 2012-2025; black = total cases, grey = suspected"

who_long |>
    filter(date >= "2020-09-01" & date <= "2025-09-01") |>
    ggplot(aes(x = date)) +
    geom_point(aes(y = measles_total)) +
    geom_line(aes(y = measles_total)) +
    geom_point(aes(y = measles_suspect), 
    colour = "grey", alpha = 0.5) +
    labs(y = NULL, x = NULL)
```

#### Vaccination coverage
