---
title: "Measles, Bangladesh 2026: descriptive epidemiology of public data"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-tools: true
    fig-width: 9
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup

library(here)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(ggridges)
library(plotly)
library(sf)
library(slider)
library(knitr)

# visualise-data palette
col_primary <- "#1a1a1a"
col_highlight <- "#c0392b"
col_reference <- "#7f7f7f"
theme_set(
  theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank(), legend.position = "bottom")
)
sf_use_s2(FALSE)

# Utility functions
# MMWR epidemiological weeks, Sunday to Saturday, labelled by start date
week_start_of <- function(date) floor_date(date, "week", week_start = 7)
week_label <- function(week) {
  paste0("week ", epiweek(week), " (", fmt_date(week), " to ",
         fmt_date(week + 6), ")")
}
# Sum daily counts by complete epiweek, grouped by the columns in ...
weekly <- function(d, ...) {
  d |>
    mutate(week = week_start_of(date)) |>
    filter(between(week, first_week, last_week)) |>
    group_by(..., week) |>
    summarise(across(c(suspected, confirmed, serum_sent), sum),
              .groups = "drop")
}
fmt <- function(x) format(round(x), big.mark = ",")
fmt_date <- function(d, year = FALSE) {
  paste(day(d), format(d, if (year) "%B %Y" else "%B"))
}
```

```{r}
#| label: inputs

# DGHS dashboard series by report date
cases <- read.csv(here("data", "dghs-cases.csv"), na.strings = c("", "NA")) |>
  mutate(date = ymd(date))
cutoff <- max(cases$date)
first_daily <- ymd("2026-04-08")
# First and last complete epiweeks of daily data
first_week <- week_start_of(first_daily - 1) + 7
last_week <- week_start_of(cutoff + 1) - 7

# DGHS district spellings that differ from the OCHA boundaries
district_recode <- c(
  Jhalakathi = "Jhalokati",
  Netrokona = "Netrakona",
  Chapainawabganj = "Chapainababganj"
)
cases <- cases |>
  mutate(district = recode(district, !!!district_recode))

# Boundaries and population (2022 census projection)
adm1 <- st_read(here("data", "covariates", "bgd_adm1.geojson"), quiet = TRUE)
adm2 <- st_read(here("data", "covariates", "bgd_adm2.geojson"), quiet = TRUE)
lookup <- read.csv(here("data", "covariates", "admin_lookup.csv"))

# Divisions west to east by centroid, and districts the same way within each
centroid_lon <- function(g) {
  suppressWarnings(st_coordinates(st_centroid(g))[, 1])
}
division_order <- adm1 |>
  mutate(lon = centroid_lon(geometry)) |>
  arrange(lon) |>
  pull(adm1_name)
district_order <- adm2 |>
  mutate(
    lon = centroid_lon(geometry),
    division = factor(adm1_name, levels = division_order)
  ) |>
  arrange(division, lon) |>
  pull(adm2_name)

pop1 <- read.csv(here("data", "covariates", "population_adm1.csv")) |>
  select(adm1_pcode, pop = t_tl, pop_u5 = t_00_04) |>
  left_join(distinct(lookup, adm1_pcode, division = adm1_name),
            by = "adm1_pcode")
pop2 <- read.csv(here("data", "covariates", "population_adm2.csv")) |>
  select(adm2_pcode, pop = t_tl, pop_u5 = t_00_04) |>
  left_join(select(lookup, adm2_pcode, district = adm2_name,
                   division = adm1_name),
            by = "adm2_pcode")
stopifnot(
  setequal(pop1$division, unique(na.omit(cases$division))),
  setequal(pop2$district, unique(na.omit(cases$district)))
)

# Wide daily series: one row per date and place, one column per measure
widen <- function(d) {
  d |>
    select(-fetched_at, -flag) |>
    pivot_wider(names_from = measure, values_from = value)
}
nat_daily <- cases |>
  filter(level == "national") |>
  widen()
div_daily <- cases |>
  filter(level == "division") |>
  widen() |>
  mutate(division = factor(division, levels = division_order))
dist_daily <- cases |>
  filter(level == "district") |>
  widen() |>
  mutate(district = factor(district, levels = district_order),
         division = factor(division, levels = division_order))
# 15 March to 7 April, loaded onto 7 April as one total
lump <- nat_daily |> filter(date == ymd("2026-04-07")) |> pull(suspected)

# Events marked on time series
events <- read.csv(here("assets", "events.csv")) |>
  mutate(date = ymd(date))
events_marked <- events |>
  filter(!grepl("backfill", event))

# WHO provisional monthly measles and rubella data
who <- read.csv(here("data", "who-monthly.csv")) |>
  mutate(month = make_date(year, month, 1)) |>
  filter(year >= 2023)
stopifnot(all(who$measles_suspect == who$measles_total + who$discarded))
```

::: {.callout-note}
# How to read this report

- Data to `r fmt_date(cutoff, year = TRUE)`, from the DGHS measles monitoring platform unless stated. Public data only.
- A case is a hospital admission with suspected measles. Counts are admissions, not infections.
- Time is the date a count was reported, a surrogate for onset.
- Place is the district of the reporting facility, not of residence.
- Weeks are MMWR epidemiological weeks, Sunday to Saturday. Weekly plots show complete weeks only, from `r week_label(first_week)` to `r week_label(last_week)`.
:::

::: {.callout-note collapse="true"}
# Data and definitions

Measles outbreak data sourced from the DGHS measles monitoring dashboard (measles.dghs.gov.bd), except where otherwise noted.
Population numbers are from the 2022 census projection.

## Data reporting process

Cases are reported as number of measles admissions to medical facilities, collated by district civil surgeon offices, with daily counts submitted to the DGHS dashboard.
Lab testing data and day of week effects in reporting are in [Appendix 2](#sec-testing).

```{r}
#| label: fig-reporting-units
#| fig-cap: >-
#|   Reporting units submitting per day, by division, Bangladesh, 8 April 2026 to cutoff.
#|   Divisions other than Dhaka: reports as a proportion of the most frequent daily count.
#|   Dhaka: count of reports.
#|   Source: DGHS measles monitoring platform.
#| fig-alt: "Line charts of daily reporting completeness by division and the daily count of reports in Dhaka division."
#| fig-height: 7

n_districts <- lookup |> count(division = adm1_name, name = "districts")
modal <- function(x) as.integer(names(which.max(table(x))))
reports <- div_daily |>
  filter(date >= first_daily) |>
  select(date, division, report_count) |>
  group_by(division) |>
  mutate(expected = modal(report_count)) |>
  ungroup() |>
  left_join(n_districts, by = "division") |>
  mutate(value = if_else(division == "Dhaka", report_count,
                         report_count / expected))
p <- ggplot(reports, aes(date, value)) +
  geom_line(colour = col_primary) +
  facet_wrap(~division, ncol = 4, scales = "free_y") +
  scale_x_date(date_labels = "%b") +
  labs(x = "Report date", y = "Reports / expected (Dhaka: count)")
ggplotly(p)

reports |>
  distinct(division, districts, expected) |>
  kable(caption = "Districts and most frequent daily count of reports, by division")
```

### Source data definitions

```{r}
#| label: tbl-measures
#| tbl-cap: "DGHS dashboard indicators, reported as 24-hour counts by report date"

tibble::tribble(
  ~measure, ~dashboard_label, ~notes,
  "suspected", "suspected24h", "Hospital admissions with suspected measles; inclusion of confirmed unknown",
  "confirmed", "confirmed24h", "Laboratory method not stated",
  "suspected_deaths", "suspectedDeath24h", "Inclusion of confirmed deaths, and place of death, unknown",
  "confirmed_deaths", "confirmedDeath24h", "As above",
  "admitted", "admitted24h", "Relation to suspected not published",
  "discharged", "discharged24h", "\"Recovered\" in the dashboard's Excel export",
  "serum_sent", "serumSent24h", "Serum samples sent to the laboratory; not in the Excel export"
) |>
  kable()
```

### WHO standard case definition

```{r}
#| label: tbl-who-definitions
#| tbl-cap: "WHO measles surveillance case definitions, used in WHO monthly data (verbatim)"

read.csv(here("assets", "case-definitions.csv")) |>
  select(classification, definition, page) |>
  kable()
```
:::

# Situation update

```{r}
#| label: situation

cum <- nat_daily |>
  summarise(across(c(suspected, confirmed, suspected_deaths,
                     confirmed_deaths), sum))
nat_weekly <- weekly(nat_daily) |>
  mutate(per_100k = 1e5 * suspected / sum(pop1$pop))
this_week <- nat_weekly |> filter(week == last_week)
prev_week <- nat_weekly |> filter(week == last_week - 7)
change_pct <- 100 * (this_week$suspected / prev_week$suspected - 1)
div_last <- weekly(div_daily, division) |>
  filter(week == last_week) |>
  left_join(pop1, by = "division") |>
  mutate(per_100k = 1e5 * suspected / pop) |>
  arrange(desc(per_100k))
dist_last <- weekly(dist_daily, district) |>
  filter(week == last_week)
partial <- nat_daily |> filter(date > last_week + 6)
partial_text <- if (nrow(partial) == 0) {
  "none; the latest week is complete"
} else {
  paste0(fmt(sum(partial$suspected)), " suspected cases in ", nrow(partial),
         " days (", fmt_date(min(partial$date)), " to ",
         fmt_date(cutoff), ")")
}
```

- Cumulative, 15 March to `r fmt_date(cutoff)`: `r fmt(cum$suspected)` suspected and `r fmt(cum$confirmed)` confirmed cases; `r fmt(cum$suspected_deaths)` suspected and `r fmt(cum$confirmed_deaths)` confirmed deaths, as published.
- Latest complete `r week_label(last_week)`: `r fmt(this_week$suspected)` suspected cases, `r sprintf("%+.0f%%", change_pct)` on the week before; `r sprintf("%.1f", this_week$per_100k)` per 100,000 population (@fig-epicurve-national).
- Highest division incidence that week: `r div_last$division[1]`, `r sprintf("%.1f", div_last$per_100k[1])` per 100,000. Districts reporting any suspected case that week: `r sum(dist_last$suspected > 0)` of 64.
- Current week to date: `r partial_text`.

<!-- TODO KS: one sentence on the current trend -->

::: {.callout-caution}
# Interpreting these data

- Counts are hospital admissions with suspected measles, not infections; cases not admitted are not counted.
- Report date stands in for onset date; the delay between them is not published ([Epidemic timespans](#sec-timespans)).
- Districts are those of the reporting facility, so referral hospitals raise counts in their own districts.
- DGHS does not publish whether suspected cases include confirmed cases, or whether suspected deaths include confirmed deaths (@tbl-measures).
- Whether cases in the Rohingya camps are counted in Cox's Bazar district is not known ([Appendix 1](#sec-camps)).
- Cases from 15 March to 7 April were reported as one total on 7 April, so daily and weekly series start on 8 April.
- WHO monthly surveillance, a separate stream, changed classification in June and disagrees with the DGHS series after May ([Appendix 3](#sec-validation)).
:::

--------------------------------------------------------------------------------

# Epidemic curve

Outbreak surveillance data appears from 15 March.
The DGHS report a single cumulative total from 15 March through 7 April (`r fmt(lump)` suspected cases), followed by daily incident data from 8 April to present.
It is likely the outbreak started substantially earlier, perhaps as early as December 2025 among the Rohingya population in Cox's Bazar.

::: {.callout-tip collapse="true"}
# Latest daily data

```{r}
#| label: latest-daily

tail(nat_daily) |> kable(caption = "Daily DGHS measles dashboard data")
```
:::

```{r}
#| label: fig-epicurve-national
#| fig-cap: >-
#|   Suspected (bars) and confirmed (line) measles cases per epidemiological week, Bangladesh, complete weeks from 12 April 2026 to the latest complete week.
#|   Confirmed cases are overlaid, not stacked: whether suspected includes confirmed is not published.
#|   Dotted lines: nationwide campaign (20 April); camp campaign and platform launch (26 April).
#|   Cases from 15 March to 7 April, reported as one total, are not shown.
#|   Source: DGHS measles monitoring platform, report date.
#| fig-alt: "Weekly bar chart of national suspected measles cases with a line of confirmed cases and dotted lines at campaign dates."

p <- nat_weekly |>
  mutate(text = paste0(week_label(week), ": ", fmt(suspected),
                       " suspected, ", fmt(confirmed), " confirmed")) |>
  ggplot(aes(week + 3.5)) +
  geom_col(aes(y = suspected, fill = "Suspected", text = text), width = 7) +
  geom_line(aes(y = confirmed, colour = "Confirmed")) +
  geom_point(aes(y = confirmed, colour = "Confirmed", text = text),
             size = 1) +
  geom_vline(data = filter(events_marked, date >= first_week),
             aes(xintercept = date), colour = col_primary,
             linetype = "dotted") +
  scale_fill_manual(values = c(Suspected = col_reference)) +
  scale_colour_manual(values = c(Confirmed = col_highlight)) +
  scale_x_date(date_labels = "%b") +
  labs(x = "Week (Sunday to Saturday)", y = "Cases per week",
       fill = NULL, colour = NULL)
ggplotly(p, tooltip = "text")
```

<!-- TODO KS: one sentence stating the pattern -->

## Epidemic timespans {#sec-timespans}

```{r}
#| label: tbl-timespans
#| tbl-cap: "Relevant epidemiological time intervals and potential data sources"

tibble::tribble(
  ~interval, ~public_data, ~line_list_variable, ~other_source,
  "Generation or serial interval", "No", "No (no transmission pairs)", "Literature",
  "Exposure to onset (incubation)", "No", "No", "Literature",
  "Fever to rash", "No", "Yes", "",
  "Rash to notification", "No", "Yes", "",
  "Notification to investigation", "No", "Yes", "",
  "Rash to specimen; specimen to result", "No", "Yes", "",
  "Onset to admission", "No", "Yes", "",
  "Admission to report date", "No; indirect signs only (days with missing reports)", "No", "",
  "Suspected to confirmed", "No (share tested varies)", "Yes", "",
  "Admission to exit (discharge or death)", "Partly: deconvolution of discharges and deaths against admissions, if deaths are hospital deaths", "Yes", "",
  "Onset to death", "No", "Free-text Comment only", "Unknown",
  "Campaign dose to rash", "No", "DateLastMCV", "",
  "Undetected period (first infection to detection)", "No", "Onset dates before April", "Molecular clock on sequences"
) |>
  kable(col.names = c("Interval", "Public data", "Line list variable",
                      "Other source"))
```

--------------------------------------------------------------------------------

# Spatial spread

```{r}
#| label: inputs-where

district_cum <- dist_daily |>
  group_by(district) |>
  summarise(suspected = sum(suspected), .groups = "drop") |>
  left_join(pop2, by = "district") |>
  mutate(per_100k = 1e5 * suspected / pop)
stopifnot(nrow(district_cum) == 64, !anyNA(district_cum$pop))

map_districts <- adm2 |>
  left_join(district_cum, by = c("adm2_name" = "district"))
breaks_100k <- c(0, 25, 50, 100, 200, 400)
stopifnot(!anyNA(map_districts$per_100k),
          max(map_districts$per_100k) <= max(breaks_100k))
```

```{r}
#| label: fig-map-cumulative
#| fig-cap: >-
#|   Cumulative suspected measles cases per 100,000 population, by district of reporting facility, Bangladesh, 15 March 2026 to cutoff.
#|   Sources: DGHS measles monitoring platform; 2022 census projection.
#| fig-alt: "Choropleth map of Bangladesh districts shaded by cumulative suspected measles cases per 100,000, with division outlines."
#| fig-height: 8

ggplot() +
  geom_sf(data = map_districts, aes(fill = per_100k), colour = "white",
          linewidth = 0.1) +
  geom_sf(data = adm1, fill = NA, colour = col_primary, linewidth = 0.4) +
  scale_fill_viridis_b(breaks = breaks_100k, limits = range(breaks_100k),
                       option = "magma", direction = -1,
                       show.limits = TRUE) +
  labs(fill = "Suspected cases\nper 100,000") +
  theme_void() +
  theme(legend.position = "right")
```

<!-- TODO KS: one sentence stating the pattern -->

```{r}
#| label: fig-ridge-district
#| fig-cap: >-
#|   Weekly suspected measles cases as a share of each district's peak week, by district of reporting facility and division, Bangladesh, complete weeks from 12 April 2026 to the latest complete week.
#|   Districts ordered within each division by week of peak, earliest at top.
#|   Heights show timing, not size: each district is scaled to its own peak.
#|   A peak in the first or last week may lie outside the series.
#|   Source: DGHS measles monitoring platform, report date.
#| fig-alt: "Ridgeline plots of weekly suspected cases scaled to each district's peak, one panel per division, districts ordered by peak week."
#| fig-height: 14

dist_weekly <- weekly(dist_daily, division, district) |>
  group_by(district) |>
  mutate(peak_week = week[which.max(suspected)]) |>
  ungroup()
# Latest peak first, so the earliest peak is the top row
ridge_order <- dist_weekly |>
  distinct(district, peak_week) |>
  arrange(desc(peak_week), desc(as.integer(district))) |>
  pull(district) |>
  as.character()

dist_weekly |>
  group_by(district) |>
  mutate(share_of_peak = suspected / max(suspected)) |>
  ungroup() |>
  ggplot(aes(week + 3.5, factor(district, levels = ridge_order),
             height = share_of_peak)) +
  geom_ridgeline(scale = 0.9, fill = col_reference, alpha = 0.5,
                 colour = col_primary, linewidth = 0.3) +
  facet_grid(division ~ ., scales = "free_y", space = "free_y") +
  scale_x_date(date_labels = "%b", expand = c(0, 0)) +
  # Room above the top ridge in each panel
  scale_y_discrete(expand = expansion(add = c(0.2, 1))) +
  labs(x = "Week (Sunday to Saturday)", y = NULL) +
  theme(axis.text.y = element_text(size = 7),
        panel.grid.major.y = element_blank(),
        strip.text.y = element_text(angle = 0))
```

<!-- TODO KS: one sentence stating the pattern -->

# Vaccination

```{r}
#| label: tbl-campaigns
#| tbl-cap: "Measles-rubella vaccination campaign rounds, Bangladesh, 2026"

events |>
  filter(grepl("campaign", event, ignore.case = TRUE)) |>
  transmute(date = fmt_date(date, year = TRUE), event, scope, source) |>
  kable(col.names = c("Date", "Event", "Scope", "Source"))
```

Not yet in the data: dates of the May round and the September mop-up, and coverage by division (`sitrep/data/dghs-campaign.csv`, not built).

--------------------------------------------------------------------------------

# Appendix 1: subpopulation outbreaks {#sec-camps}

## Rohingya population

```{r}
#| label: inputs-camps

# EWARS weekly suspected measles/rubella in the camps, digitised from the
# W36 2026 bulletin chart; confirmed as a series (data/camps-review.csv)
ewars <- read.csv(here("data", "camps-ewars-weekly.csv")) |>
  filter(confirmed_by != "") |>
  transmute(week = ymd(week_start), ewars = value)
# DGHS admissions reported by facilities in Cox's Bazar district
coxs_weekly <- dist_daily |>
  filter(district == "Cox's Bazar") |>
  weekly() |>
  select(week, dghs = suspected)
camps_weekly <- full_join(ewars, coxs_weekly, by = "week") |>
  arrange(week)

# Camp case and death figures from situation reports and bulletins,
# confirmed against the source page
camps_figures <- read.csv(here("data", "camps-measles.csv"),
                          na.strings = "") |>
  filter(!is.na(confirmed_by),
         indicator %in% c("camp_suspected_cases", "camp_confirmed_cases",
                          "camp_epilinked_cases", "camp_total_cases",
                          "camp_deaths", "camp_admissions")) |>
  mutate(date = ymd(coalesce(as_of_date, published_date)))
```

```{r}
#| label: fig-epicurve-camps
#| fig-cap: >-
#|   Suspected measles/rubella cases per epidemiological week in the Rohingya camps (bars), and suspected measles admissions reported by facilities in Cox's Bazar district (line), 22 December 2024 to 12 September 2026.
#|   Camp counts are syndromic reports from camp health facilities, not hospital admissions; the two series are not comparable in kind.
#|   Whether camp cases are included in Cox's Bazar district's DGHS count is not known.
#|   Dotted line: camp MR campaign (26 April).
#|   Sources: Rohingya Response EWARS bulletin W36 2026 (Figure 4, digitised); DGHS measles monitoring platform, report date.
#| fig-alt: "Weekly bar chart of suspected measles cases in the Rohingya camps from late 2024, with a line of weekly suspected admissions in Cox's Bazar district from April 2026 and a dotted line at the camp campaign."

p <- camps_weekly |>
  mutate(text = paste0(week_label(week), ": ",
                       if_else(is.na(ewars), "", paste0(fmt(ewars), " camps")),
                       if_else(is.na(dghs), "",
                               paste0("; ", fmt(dghs), " Cox's Bazar (DGHS)")))) |>
  ggplot(aes(week + 3.5)) +
  geom_col(aes(y = ewars, fill = "Camps (EWARS)", text = text), width = 7,
           na.rm = TRUE) +
  geom_line(aes(y = dghs, colour = "Cox's Bazar district (DGHS)"),
            na.rm = TRUE) +
  geom_point(aes(y = dghs, colour = "Cox's Bazar district (DGHS)",
                 text = text), size = 1, na.rm = TRUE) +
  geom_vline(data = filter(events, scope == "camps"),
             aes(xintercept = date), colour = col_primary,
             linetype = "dotted") +
  scale_fill_manual(values = c("Camps (EWARS)" = col_reference)) +
  scale_colour_manual(values = c("Cox's Bazar district (DGHS)" =
                                   col_highlight)) +
  scale_x_date(date_labels = "%b %Y") +
  labs(x = "Week (Sunday to Saturday)", y = "Suspected cases per week",
       fill = NULL, colour = NULL)
ggplotly(p, tooltip = "text")
```

<!-- TODO KS: one sentence stating the pattern -->

::: {.callout-tip collapse="true"}
# Camp case and death figures in situation reports

```{r}
#| label: tbl-camps-figures
#| tbl-cap: >-
#|   Camp measles case and death figures as stated in situation reports and bulletins, each confirmed against its source page.
#|   Geography, period and case base differ between documents and are given as stated.
#|   Sources: Health Sector Cox's Bazar bulletins; UN RCO, UNICEF and WHO SEARO reports.

camps_figures |>
  arrange(indicator, date) |>
  transmute(
    measure = sub("^camp_", "", indicator) |> gsub(pattern = "_", replacement = " "),
    value = fmt(value),
    date = if_else(is.na(date), "", format(date, "%e %b")),
    geography, period = coalesce(period_text, period, ""),
    source = doc_id, note = coalesce(gsub("\n", " ", review_note), "")
  ) |>
  kable(col.names = c("Measure", "Value", "As of or published", "Geography",
                      "Period", "Document", "Reviewer note"))
```
:::

--------------------------------------------------------------------------------

# Appendix 2: Test positivity and data reporting pipeline {#sec-testing}

## Lab testing

Serum samples are recorded on the platform from May.
Serum sent per suspected case and confirmed per suspected case are ratios of totals since 1 May.
Neither is test positivity: the platform does not report negative results, and a confirmation need not come from a sample recorded on the platform.

```{r}
#| label: tbl-testing
#| tbl-cap: >-
#|   Suspected cases, serum samples sent and confirmed cases, by division, Bangladesh, 1 May 2026 to cutoff.
#|   Source: DGHS measles monitoring platform, report date.

serum_start <- ymd("2026-05-01")
sum_testing <- function(d) {
  d |>
    filter(date >= serum_start) |>
    summarise(across(c(suspected, serum_sent, confirmed), sum))
}
bind_rows(
  div_daily |>
    group_by(place = as.character(division)) |>
    sum_testing(),
  nat_daily |>
    sum_testing() |>
    mutate(place = "National")
) |>
  mutate(serum_per_suspected = serum_sent / suspected,
         confirmed_per_suspected = confirmed / suspected) |>
  kable(digits = 2, format.args = list(big.mark = ","),
        col.names = c("Place", "Suspected", "Serum sent", "Confirmed",
                      "Serum sent per suspected",
                      "Confirmed per suspected"))
```

WHO monthly surveillance records discarded cases, so it gives the only laboratory positivity in public data.
Share tested = (laboratory-confirmed + discarded) / suspected; an upper bound, since cases can be discarded without a test (@tbl-who-definitions).
Laboratory positivity = laboratory-confirmed / (laboratory-confirmed + discarded).

```{r}
#| label: tbl-who-positivity
#| tbl-cap: >-
#|   Measles cases by final classification, share tested and laboratory positivity, by month, Bangladesh, 2026.
#|   Source: WHO provisional monthly measles and rubella data.

who |>
  filter(year == 2026) |>
  transmute(
    month = format(month, "%B"),
    suspected = measles_suspect,
    lab_confirmed = measles_lab_confirmed,
    epi_linked = measles_epi_linked,
    clinical = measles_clinical,
    discarded,
    share_tested = (measles_lab_confirmed + discarded) / measles_suspect,
    positivity = measles_lab_confirmed / (measles_lab_confirmed + discarded)
  ) |>
  kable(digits = 2, format.args = list(big.mark = ","),
        col.names = c("Month", "Suspected", "Lab-confirmed", "Epi-linked",
                      "Clinically compatible", "Discarded", "Share tested",
                      "Lab positivity"))
```

## Day of week reporting

```{r}
#| label: tbl-weekday
#| tbl-cap: >-
#|   Daily suspected cases relative to the centred 7-day mean, by weekday and division: median (interquartile range), Bangladesh, 8 April 2026 to cutoff.
#|   Source: DGHS measles monitoring platform, report date.

weekday <- div_daily |>
  filter(date >= first_daily) |>
  arrange(division, date) |>
  group_by(division) |>
  mutate(centred = slide_dbl(suspected, mean, .before = 3, .after = 3,
                             .complete = TRUE),
         rel = suspected / centred) |>
  ungroup() |>
  filter(!is.na(rel)) |>
  mutate(weekday = wday(date, label = TRUE, week_start = 6)) |>
  group_by(division, weekday) |>
  summarise(cell = sprintf("%.2f (%.2f to %.2f)", median(rel),
                           quantile(rel, 0.25), quantile(rel, 0.75)),
            .groups = "drop") |>
  pivot_wider(names_from = weekday, values_from = cell)
kable(weekday)
```

# Appendix 3: data validation {#sec-validation}

## Comparison with WHO

WHO still produce monthly measles and rubella surveillance data, but with an apparent change in case ascertainment from June.

```{r}
#| label: fig-timeline
#| fig-cap: >-
#|   Measles cases per month: WHO surveillance suspected and total cases (bars), and DGHS hospital suspected cases (points), Bangladesh, January 2026 to cutoff.
#|   DGHS April is 8 to 30 April; 15 March to 7 April is reported as one total; the current month is shown when complete.
#|   Sources: WHO provisional monthly measles and rubella data; DGHS measles monitoring platform.
#| fig-alt: "Monthly bar chart of WHO suspected and total measles cases with DGHS monthly suspected hospital cases as points."

who_2026 <- who |>
  filter(year(month) == 2026) |>
  select(month, `WHO suspected` = measles_suspect,
         `WHO total (lab + epi-linked + clinical)` = measles_total) |>
  pivot_longer(-month, names_to = "series", values_to = "cases")
dghs_monthly <- nat_daily |>
  filter(date >= first_daily) |>
  group_by(month = floor_date(date, "month")) |>
  summarise(cases = sum(suspected), days = n(), .groups = "drop") |>
  # Current month dropped while incomplete; April starts on 8 April
  filter(month < floor_date(cutoff + 1, "month")) |>
  mutate(series = "DGHS suspected (hospital)")

p <- ggplot() +
  geom_col(data = who_2026,
           aes(month, cases, fill = series),
           position = position_dodge(width = 25), width = 24) +
  geom_point(data = dghs_monthly,
             aes(month, cases, colour = series, text = paste(days, "days")),
             size = 2.5) +
  geom_vline(xintercept = ymd("2026-06-01"), linetype = "dotted",
             colour = col_reference) +
  annotate("text", x = ymd("2026-06-01"), y = max(dghs_monthly$cases),
           label = "WHO classification change", hjust = -0.05,
           size = 3, colour = col_reference) +
  scale_fill_manual(values = c(
    `WHO suspected` = "#d4d4d4",
    `WHO total (lab + epi-linked + clinical)` = col_reference)) +
  scale_colour_manual(values = c(`DGHS suspected (hospital)` = col_highlight)) +
  scale_x_date(date_labels = "%b") +
  labs(x = NULL, y = "Cases per month", fill = NULL, colour = NULL)
ggplotly(p, tooltip = c("x", "y", "text"))
```

## Data quality

```{r}
#| label: tbl-data-quality
#| tbl-cap: "Failed checks and flagged dates, DGHS dashboard series, Bangladesh, 2026"

checks <- read.csv(here("data", "dghs-cases-checks.csv"))
failed <- checks |>
  filter(!pass) |>
  transmute(item = check, date, division, measure,
            detail = paste("expected", expected, "observed", observed))
flagged <- cases |>
  filter(!is.na(flag), level == "national", measure == "suspected") |>
  transmute(item = flag, date = as.character(date), division = NA,
            measure = NA, detail = paste("national suspected", value))
bind_rows(failed, flagged) |> kable()
```

```{r}
#| label: tbl-source-agreement
#| tbl-cap: "Agreement of dashboard and press release counts by month, Bangladesh, 2026"

read.csv(here("sitrep", "data", "dghs-source-compare.csv")) |>
  mutate(month = format(ymd(date), "%Y-%m")) |>
  filter(!is.na(diff)) |>
  group_by(month) |>
  summarise(comparisons = n(), differ = sum(diff != 0),
            net_difference = sum(diff)) |>
  kable()
```
